Negative regulation of APC/C activation by MAPK-mediated attenuation of Cdc20Slp1 under stress

Version 1

Abstract

Abstract

Mitotic anaphase onset is a key cellular process tightly regulated by multiple kinases. The involvement of mitogen-activated protein kinases (MAPKs) in this process has been established in Xenopus egg extracts. However, the detailed regulatory cascade remains elusive, and also it is unknown whether the MAPKs-dependent mitotic regulation is evolutionarily conserved in the single cell eukaryotic organism such as fission yeast (Schizosaccharomyces pombe). Here we show that two MAPKs in S. pombe indeed act in concert to restrain anaphase-promoting complex/cyclosome (APC/C) activity upon activation of the spindle assembly checkpoint (SAC). One MAPK, Pmk1, binds and phosphorylates Slp1Cdc20, the co-activator of APC/C. Phosphorylation of Slp1Cdc20 by Pmk1, but not by Cdk1, promotes its subsequent ubiquitylation and degradation. Intriguingly, Pmk1-mediated phosphorylation event is also required to sustain SAC under environmental stress. Thus, our study establishes a new underlying molecular mechanism of negative regulation of APC/C by MAPK upon stress stimuli, and provides an unappreciated framework for regulation of anaphase entry in eukaryotic cells.

Introduction

The evolutionarily conserved mitogen-activated protein kinase (MAPK) signaling pathways regulate multiple cellular functions in eukaryotic organisms in response to a wide variety of environmental cues (Plotnikov et al., 2011). However, different MAPK pathways have been evolved in an organism to integrate diverse signals and fulfill different regulations on various effectors (Cansado et al., 2021; Ronkina and Gaestel, 2022). This makes specifying the MAPKs and substrates involved in a specific physiological process rather complex and challenging.

In late mitosis, the timely polyubiquitylation and subsequent degradation of securin and cyclin B by anaphase-promoting complex/cyclosome (APC/C) play a critical role for anaphase onset and chromosome segregation (Peters, 2006; Sullivan and Morgan, 2007; Yamano, 2019). Given the essential role of APC/C in triggering chromosome segregation, it is not surprising that APC/C is a key molecular target of the spindle assembly checkpoint (SAC), which is an intricate surveillance mechanism that prolongs mitosis until all chromosomes achieve correct bipolar attachments to spindle microtubules (McAinsh and Kops, 2023; Murray, 2011; Musacchio, 2015). Previous studies have revealed that the active MAPK is required for spindle checkpoint activation and APC/C inhibition in Xenopus egg extracts or tadpole cells (Chen, 2004; Chung and Chen, 2003; Minshull et al., 1994; Takenaka et al., 1997; Zhao and Chen, 2006). These studies suggested that phosphorylation of several components of SAC, including Cdc20, Mps1 and Bub1, by MAPK serves as regulatory signals to block APC/C activation upon SAC activation (Chen, 2004; Chung and Chen, 2003; Zhao and Chen, 2006). However, all these studies used either the anti-MAPK antibodies or small molecule inhibitor UO126 to inhibit MAPK or MAPK-specific phosphatase MKP-1 to antagonize MAPK activity, these reagents could not distinguish the contributions of different MAPKs with high specificity. Thus, those seemingly defined mechanisms might suffer the ambiguity due to potential specificity issues.

The fission yeast Schizosaccharomyces pombe has three MAPK-signaling cascades: the pheromone signaling pathway (PSP), the stress-activated pathway (SAP) and the cell integrity pathway (CIP), with Spk1, Sty1 or Pmk1 as MAPK respectively (Figure 1A) (Cansado et al., 2021; Perez and Cansado, 2010). In S. pombe, whether MAPKs play key roles in regulation of the APC/C and spindle checkpoint remains largely unexplored. Intriguingly, all genes encoding the major components of three MAPK-signaling cascades in fission yeast are non-essential (Kim et al., 2010), this provides the possibility for investigations on their potential functions in a genetic background with complete deletion of relevant genes.

pmk1Δ and sty1-T97A mutants display spindle checkpoint activation defects, while pek1DD and wis1DD mutants are defective in checkpoint silencing.

(A) Schematic of core modules of three Schizosaccharomyces pombe mitogen-activated protein kinase (MAPK) signaling pathways. Each cascade consists of three core kinases (MAP kinase kinase kinase (MAPKKK), MAP kinase kinase (MAPKK), and MAPK). CIP, the cell integrity pathway; SAP, the stress-activated pathway; PSP, pheromone signaling pathway. These pathways can be disrupted by pmk1Δ, sty1Δ or ATP analogue-sensitive mutation sty1-T97A or spk1Δ, and constitutively activated by mutations in MAPKKs Pek1 (pek1DD, pek1-S234D;T238D) or Wis1 (wis1DD, wis1-S469D;T473D) respectively.

(B) Serial dilution assay on TBZ sensitivity of all MAPK deletion mutants and pek1DD- or wis1DD-overexpressing mutants. mad2Δ is a positive control. Note Padh11 is a stronger version of Padh21 promoter.

(C) Schematic depiction of the experiment design for time-course analyses on SAC or APC/C activation. nda3-KM311 cells carrying Cdc13-GFP were grown, synchronized with HU and treated at 18 °C for 6 hours to activate SAC, and finally shifted back to the permissive temperature 30 °C. Samples were collected at 10 min intervals and subjected to microscopy analyses. Example pictures of cells with Cdc13-GFP signals enriched or disappeared at spindle pole bodies (SPBs) are shown. Scale bar, 5 μm.

(D) Time-course analyses of SAC activation and inactivation in nda3-KM311 cdc13-GFP strains with indicated genotypes. sty1-T97A was inactivated by 5μM 3-BrB-PP1.

In this work, we set out to directly examine the requirement of the fission yeast MAPKs in mitotic APC/C activation and anaphase entry. Our results uncovered a previously unappreciated mechanism in which Pmk1, the MAPK of CIP, phosphorylates the APC/C activator Slp1 (the Cdc20 homologue in S. pombe) upon SAC activation. Phosphorylation of Slp1Cdc20 results in its ubiquitylation and degradation and thus subsequently impedes APC/C activation and anaphase entry. This mechanism also operates in response to spindle defects under environmental stress. Therefore, our finding extends cell-cycle control stages linked to MAPKs from previously recognized G2/M transition and cytokinesis to anaphase entry and mitotic exit in fission yeast. It will also be interesting to address whether this MAPK-mediated Cdc20 regulation exists as a general mechanism for balancing spindle checkpoint activity in higher eukaryotes.

Results

SAP and CIP signaling components are required for spindle checkpoint activity

As the first step to examine the possible requirement of the fission yeast MAPKs in mitotic progression, we tested the sensitivity of deletion mutants of three fission yeast MAPKs to microtubule-destabilizing drug thiabendazole (TBZ), which has been routinely used as an indicator of defective kinetochore or spindle checkpoint (Akera et al., 2015; Saitoh et al., 1997). We noticed that pmk1Δ and sty1Δ but not spk1Δ cells were extremely sensitive to TBZ, and constitutively activated CIP signaling by overexpressing Pek1DD (Pek1-S234D;T238D) (Sugiura et al., 1999) or activated SAP signaling by overexpressing Wis1DD (Wis1-S469D;T473D) (Shiozaki et al., 1998) also caused yeast cells to be sensitive to TBZ (Figure 1B; Supplementary Figure 1A). These observations indicated that Pmk1 and Sty1 pathways might be involved in mitosis-related processes.

To extend our analysis and precisely quantify whether the SAC is properly activated and maintained in MAPK mutants, we adopted one well-established assay using the cold-sensitive β-tubulin mutant nda3-KM311, which compromises the kinetochore-spindle microtubule attachment upon cold treatment (Hiraoka et al., 1984), to analyze the ability of cells to arrest in mitosis in the absence of spindle microtubules and enter anaphase when spindle microtubules are reassembled (Bai et al., 2022; May et al., 2017; Vanoosthuyse and Hardwick, 2009). In this assay, the SAC was first robustly activated by the nda3-KM311 mutant at 18 °C and then inactivated simply by shift mitotically arrested cells back to permissive temperature (30 °C) (Figures 1C). Because Cdc13 (cyclin B in S. pombe) localizes to the spindle pole bodies (SPBs) in early mitosis and should be degraded by APC/C to promote metaphase-anaphase transition, accumulation of Cdc13-GFP on SPBs after cold treatment serves as the read-out of the SAC activation, and the disappearance rate of Cdc13-GFP spot under permissive temperature reflects the SAC inactivation efficacy (Figures 1C). It is noteworthy that the ATP analogue-sensitive mutant sty1-T97A (i.e. sty1-as2) (Zuin et al., 2010), instead of sty1Δ deletion mutant, was used in this assay. That was due to the G2 delay that has been demonstrated in the sty1Δ deletion mutant (Shiozaki and Russell, 1995a), which could interfere mitotic timing for further analysis in this assay. We found that inactivation of CIP or SAP signaling by pmk1Δ or sty1-T97A compromised full SAC activation, as these two mutants had only less than 60% of cells with Cdc13-GFP on SPBs after 6 hours of cold treatment, whereas the percentage in wild-type and spk1Δ cells after the same treatment was roughly 90% (Figure 1D; Supplementary Figure 1C). Consistently, the disappearance of Cdc13-GFP from SPBs in pek1DD and wis1DD mutant cells released from metaphase-arrest was severely delayed (Figure 1D; Supplementary Figure 1B), suggesting delayed SAC inactivation and anaphase onset after release from spindle checkpoint arrest in these mutants.

All above data suggested that MAPK pathways are evolutionarily conserved in the fission yeast to participate in spindle checkpoint activation and maintenance, of which the CIP and SAP signaling pathways, but not the Spk1 pathway, are actively involved.

CIP and SAP signaling attenuate protein levels of Slp1Cdc20 and facilitate association of MCC with APC/C respectively

The mitotic checkpoint complex (MCC, consisting of Mad3-Mad2-Slp1Cdc20 in S. pombe) is the most potent inhibitor of the APC/C prior to anaphase (Kapanidou et al., 2017; Primorac and Musacchio, 2013; Sudakin et al., 2001). Previous study has revealed that Xenopus MAPK phosphorylates Cdc20 and facilitates its binding by Mad2, BubR1 (Mad3 in S. pombe) and Bub3 to form a tight MCC to prevent Cdc20 from activating the APC/C (Chung and Chen, 2003). In fission yeast, the key SAC components Mad2 and Mad3 and one molecule of Slp1Cdc20 form MCC which binds to APC/C through another molecule of Slp1Cdc20 upon checkpoint arrest, and the recovery from mitotic arrest accompanies the loss of MCC-APC/C binding (May et al., 2017; Sczaniecka et al., 2008; Sewart and Hauf, 2017; Vanoosthuyse and Hardwick, 2009). Since we found that constitutive activation of CIP or SAP signaling in fission yeast by pek1DDor wis1DD mutations caused prolonged SAC activation, we were suspicious that these two MAPK pathways might also execute their functions cooperatively through a similar mechanism to that in Xenopus.

To test this possibility, we analyzed the MCC occupancy on APC/C by immunoprecipitations of the APC/C subunit Lid1 (Yoon et al., 2002) in pek1DDand wis1DD cells arrested by activated checkpoint. Compared to wild-type, more MCC components (Mad2 plus Mad3 and Slp1Cdc20) were co-immunoprecipitated in wis1DD cells, while less of them were co-immunoprecipitated in pek1DD cells (Figure 2A), suggesting that activation of the SAP but not the CIP pathway enhanced the association of MCC with APC/C.

Upon spindle checkpoint activation, Slp1Cdc20 levels are reduced in pek1DD and the MCC-APC/C association is enhanced in wis1DD cells compared to wild-type cells.

(A) Co-immunoprecipitation analysis on MCC-APC/C association. Cells with indicated genotypes were grown at 30 °C to mid-log phase and arrested at 18 °C for 6 hours. Lid1-TAP was immunoprecipitated and associated Mad2, Mad3 and Slp1Cdc20 were detected by immunoblotting. The amount of co-immunoprecipitated Mad2, Mad3 and Slp1Cdc20 was quantified by being normalized to those of total immunoprecipitated Lid1 in each sample, with the relative ratio between Mad2-GFP plus Mad3-GFP or Slp1Cdc20 and Lid1-TAP in wild-type sample set as 1.0. Blots are representative of three independent experiments. p values were calculated against wild-type cells. *, p<0.05; **, p<0.01; ***, p<0.001; ****, p<0.0001; n.s., not significant.

(B, C) Immunoblot analysis of Slp1Cdc20 abundance in nda3-KM311 cells treated at 18 °C for 6 hr. Slp1Cdc20 levels were quantified with the relative ratio between Slp1Cdc20 and Cdc2 in wild-type strain set as 1.0. Phosphorylated Pmk1 (Pmk1-P) or phosphorylated Sty1 (Sty1-P) in (A) were detected using anti-phospho p42/44 and anti-phospho p38 antibodies and represents activated CIP or SAP signaling, respectively. sty1-T97A was inactivated by 5μM 3-BrB-PP1. Blots shown are the representative of three independent experiments. p values were calculated against wild-type cells. ***, p<0.001; n.s., not significant.

(D) RT-qPCR analysis of mRNA levels of slp1+. Cells with indicated genotypes were grown and treated as in (A-C) before RNA extraction. The relative fold-change (slp1+/act1+) in mRNA expression was calculated with that in wild-type cells being normalized to 1.0. Note mRNA level of slp1+ in pek1DD mutant is not decreased. Error bars indicate mean ±standard deviation of three independent experiments. Two-tailed unpaired t-test was used to derive p values. n.s, not significant.

(E) Schematic summary of the negative effect of activated CIP and SAP signaling on APC/C activation based on primary phenotype characterization of pmk1Δ, sty1-T97A, pek1DDand wis1DD mutants.

The decrease of co-immunoprecipitated Mad2 and Mad3 in pek1DDcells was largely out of expectation. Further analysis of the inputs of these immunoprecipitations showed that Slp1Cdc20 levels were significantly reduced in pek1DD but not wis1DD cells (Figure 2A). Our immunoblotting analyses using the total lysate of pek1DD or wis1DD cells arrested at metaphase also confirmed these results (Figure 2B). Consistently, the Slp1Cdc20 level was significantly increased in pmk1Δ and pmk1Δ sty1-T79A cells but not in sty1-T79A cells (Figure 2C). The elevated Slp1Cdc20 level in pmk1Δ cells is similar to that we observed in cells harboring two copies of slp1+, which leads to artificially increased Slp1Cdc20 level and also causes a SAC strength defect (Bai et al., 2022). It is noteworthy that attenuated levels of Slp1Cdc20 in pek1DD cells was not due to decreased transcription of slp1+ gene (Figure 2D). These results collectively revealed a negative effect of CIP but not SAP signaling on Slp1Cdc20 levels, which potentially results in the weakened MCC-APC/C interaction detected in CIP mutants (Figure 2A).

Together, our above data suggested a new mechanism involves division of labor between the two fission yeast MAPK pathways CIP (Pek1-Pmk1) and SAP (Wis1-Sty1) upon SAC activation, which is required in concert to lower Slp1Cdc20 levels and enhance MCC affinity for APC/C to strongly inhibit APC/C activity (Figure 2E). In this study, we only focused on characterizing the role of the CIP/Pmk1 in regulating SAC inactivation and anaphase onset hereafter.

Pmk1 directly binds Slp1Cdc20 to attenuate its levels upon SAC activation

To gain further insight into how Slp1Cdc20 levels are modulated by Pmk1, we examined the potential interaction of Pmk1 with Slp1Cdc20. Through incubating MBP-tagged Slp1Cdc20 with metaphase-arrested yeast cell lysates, we found that Pmk1 expressed from yeast cells could be pull-downed by Slp1Cdc20 (Figure 3A).

Pmk1 directly binds Slp1Cdc20 to attenuate its levels upon spindle checkpoint activation.

(A) In vitro binding assay using bacterially expressed MBP-Slp1Cdc20 and yeast lysates prepared from nda3-KM311 pmk1-HA-6His cells arrested at 18 °C for 6 hr. Note that weak band detected in MBP sample was due to unspecific background binding to amylose beads. Coomassie blue staining shows inputs for MBP and MBP-Slp1Cdc20.

(B) Schematic depiction of the S. pombe Slp1Cdc20 protein structure with the positions of two confirmed basic-residue patches mediating Slp1Cdc20-Pmk1 association indicated by pink bars. Alignment highlights the conservation of basic-residue patches within 4 Schizosaccharomyces species. The deduced Pmk1-docking motifs and different versions of motif mutations are shown. MIM, Mad2-interaction motif; IR, isoleucine-arginine tail.

(C) In vitro GST pull-down assays with bacterially expressed recombinant GST-Pmk1 and MBP-fusions of wild-type Slp1Cdc20 or Slp1Cdc20 mutants harboring Pmk1-docking motif mutations. An aliquot of the same amount of MBP-Slp1Cdc20 as that added in each GST pull-down reactions was immobilized by amylose resin as the input control. Asterisks indicate unspecific or degraded protein bands.

(D) Immunoblot analysis of Slp1Cdc20 abundance in nda3-KM311 cells with indicated genotypes treated at 18 °C for 6 hr. Slp1Cdc20 levels were quantified as in Figure 2B and 2C. The experiment was repeated 3 times and the mean value for each sample was calculated. *, p<0.05; **, p<0.01; ***, p<0.001; ****, p<0.0001; n.s., not significant.

(E) Time-course analyses of SAC activation and inactivation in nda3-KM311 cdc13-GFP strains with indicated genotypes. The experiment was repeated 3 times and the mean value for each sample was calculated as in Figure 1D.

(F) Schematic summarizing the negative effect of Pmk1-Slp1Cdc20 association on Slp1Cdc20 abundance and APC/C activation.

In previous reports, MAPK-docking sites, which are important for interaction with MAPKs, have been identified at the N termini of different MAPKKs from yeast to humans with a notable feature of a cluster of at least two basic residues (mainly Lys/K and Arg/R) separated by a spacer of 2-6 residues from a hydrophobic-X-hydrophobic sequence (Bardwell et al., 2001; Bardwell and Thorner, 1996). Thus, we attempted to identify the potential key residues in Slp1Cdc20 that mediate the Pmk1-Slp1Cdc20 interaction. By visual scanning of Slp1Cdc20 sequence, we noticed five basic-residue patches which are also present in several fungal species and loosely resemble the consensus sequence of the known MAPK-docking sites, and four of them are within N-terminal portion of the protein (Figure 3B and Supplementary Figure 2). However, only mutating either of the two most N-terminal basic-residue patches (19-KKR-21 and 47-KR-48) to glutamates (Glu, E) or alanines (Ala, A) in Slp1Cdc20 resulted in decreased amount of Slp1Cdc20 pulled down by Pmk1-GST in vitro using the bacterially expressed recombinant proteins, and mutating both patches or mutating only two arginine residues (R21 and R48) in those patches was sufficient to completely disrupt their interaction (Figure 3C and Supplementary Figure 3). Importantly, mutations of these Pmk1-docking sequences in Slp1 (PDSS for short) to alanines mimicked pmk1Δ mutant in both elevated Slp1Cdc20 protein levels (Figure 3D) and defective activation and maintenance of the SAC upon nda3-mediated checkpoint arrest (Figure 3E and Supplementary Figure 4). Also, similar to removal of Pek1 downstream MAPK Pmk1, the lowered Slp1Cdc20 protein levels and delayed SAC inactivation in pek1DD mutant could be relieved by slp1(DS-5A) mutant, which is deficient in Pmk1-docking (Figure 3D, 3E and Supplementary Figure 4).

Previously, one conserved non-canonical Pmk1-docking sequence which bears positively charged residues and IYT motif has been identified in MAPK phosphatase Pmp1 (Sacristan-Reviriego et al., 2014). Very strikingly, insertion of a 15 amino acid sequence containing the Pmk1-docking sequence in Pmp1 (or PDSP for short) right in front of the first basic-residue patch in Slp1(DS-5A) mutant protein restored its interaction with Pmk1 (Figure 3B and 3C). And this artificially forced Pmk1-Slp1Cdc20 interaction via PDSP was able to lower Slp1Cdc20 protein levels and restore higher percentage of metaphase arrest upon SAC activation in slp1(DS-5A) mutant, which is Pmk1-docking-deficient, although the rescuing effect was less efficient than the presence of genuine PDSS (Figure 3D, 3E and Supplementary Figure 4). These data firmly supported the idea that Pmk1 binds Slp1Cdc20 to downregulate its levels upon SAC activation.

Collectively, these results demonstrated that Pmk1 directly binds to Slp1Cdc20 through two short N-terminal basic patches serving as the major docking sites for Pmk1, and the physical interaction of Pmk1-Slp1Cdc20 is involved in attenuation of Slp1Cdc20 abundance upon SAC activation (Figure 3F).

Pmk1 synergizes with Cdk1 to phosphorylate Slp1Cdc20

The association of Pmk1 with Slp1Cdc20 and the negative effect of constitutively activated CIP signaling on SAC inactivation relies on Pmk1 kinase suggested that Slp1Cdc20 may be a direct substrate of Pmk1. To test this possibility, we performed in vitro phosphorylation reactions and found that bacterially expressed recombinant GST-Slp1Cdc20 could be efficiently phosphorylated in the presence of Pmk1-HA-His purified from yeast cells (Figure 4A).

Pmk1 synergizes with Cdk1 to phosphorylate and reduce Slp1Cdc20 abundance.

(A) Non-radioactive in vitro phosphorylation assays with bacterially expressed recombinant GST-Slp1Cdc20 and Pmk1-HA-His purified from yeast cells. The incorporation of the thiophosphate group was determined using anti-thiophosphate ester antibodies (anti-thioP) as indicative of phosphorylation. Note that the presence or absence of GST-Pek1DD does not affect Slp1Cdc20 phosphorylation efficiency. A known Pmk1 substrate Atf1 was used as a positive control (Atf1-P). Asterisks indicate bands corresponding to unspecific or likely degraded proteins.

(B) Summary of mass spectrometry data on Slp1Cdc20 phosphorylation and ubiquitylation in vivo in nda3-KM311-arrested cells. Red arrows and filled blue circles denote all detected phosphorylated or ubiquitylated sites respectively, while gray arrows and unfilled circles indicate the absence of phosphorylation or ubiquitylation in some of these sites respectively. Alignment highlights the conservation of most of the detected phosphorylation and ubiquitylation sites within 4 Schizosaccharomyces species.

(C) In vitro phosphorylation assays with bacterially expressed recombinant GST-fusion of Slp1Cdc20 fragment (456-488aa), GST-Pmk1 and GST-Pek1DD. The reactions were blotted with pT480 antibodies. Asterisks indicate bands corresponding to unspecific or likely degraded proteins.

(D) Immunoblot detection of Slp1Cdc20 phosphorylation at T480 in vivo. GST-slp1(456-488aa) was purified from mts3-1 cells with indicated genotypes arrested at 36℃ for 3.5 hours, and detected with anti-GST and anti-pThr480 antibodies. Note that pThr480 is absent in pmk1Δ cells, and enhanced in pek1DDcells relative to that in wild type cells.

(E) In vitro phosphorylation assays with bacterially expressed recombinant MBP-fusion of Slp1Cdc20 fragment (1-190aa) and Cdc13 (cyclin B)-containing Cdk1 complexes purified from metaphase-arrested nda3-KM311 yeast cells. 1-NM-PP1 was added as inhibitor for analogue-sensitive Cdc2-as. The reactions were blotted with pS28/pT31 antibodies. Asterisks indicate bands corresponding to unspecific or likely degraded proteins.

(F) Immunoblot detection of Slp1Cdc20 phosphorylation at S28/T31 in vivo. Apc15-13myc was immunoprecipitated from nda3-KM311 cells treated at 18 °C for 6 hr and the samples were blotted with pS28/pT31 antibodies. One IP sample from wild type background was treated with λ-phosphatase. For cdc2-asM17 cells, 1-NM-PP1 was added to inactivate Cdc2 during culturing.

(G) Immunoblot analysis of Slp1Cdc20 abundance in nda3-KM311 cells treated at 18 °C for 6 hr. Slp1Cdc20 levels were quantified as in Figure 2B and 2C. The experiment was repeated 3 times and the mean value for each sample was calculated. *, p<0.05; **, p<0.01; ***, p<0.001; ****, p<0.0001; n.s., not significant.

(H) Time-course analyses of SAC activation and inactivation in nda3-KM311 cdc13-GFP strains with indicated genotypes. The experiment was repeated 3 times and the mean value for each sample was calculated as in Figure 1D.

(I) Schematic summarizing the negative effect of Slp1Cdc20 phosphorylation by Pmk1 on its abundance and APC/C activation.

To identify potential phosphorylation sites in vivo, we purified Apc15-GFP from wild type, pmk1Δ and pek1DD cells arrested by activated checkpoint with intention that the Slp1Cdc20-containing APC/C complexes could be captured (Supplementary Figure 5A and 5B). Purifications were performed in the presence of phosphatase inhibitors to enrich for phosphopeptides. Mass spectrometry analysis revealed that the phosphorylation of Thr480 residue at the most C-terminus of Slp1Cdc20 was dependent on the presence of Pmk1, and this modification was present in pek1DD but not in wild type cells (Figure 4B, Supplementary Figure 5C-E and Supplementary Figure 6), indicating that phosphorylation of Thr480 is very likely triggered by CIP signaling. We raised phospho-specific antibodies against pT480, it could recognize Slp1Cdc20 fragment phosphorylated by Pmk1 in vitro and mutation of Thr480 to alanine (T480A) abolished site-specific phosphorylation (Figure 4C). We also confirmed that endogenous Slp1Cdc20 was indeed phosphorylated at Thr480 in metaphase-arrested mts3-1 cells, which was enhanced in pek1DD mutant and diminished in pmk1Δ mutant (Figure 4D). Thus, our results established Thr480 as a major Pmk1 phosphorylation site in vivo upon SAC activation.

In addition to pT480, our mass spectrometry analysis also identified Ser28, Thr31 and Ser59 as Pmk1-independent and Ser76 as Pmk1-dependent phosphorylation sites (Figure 4B, Supplementary Figure 5C-E and Supplementary Figure 6). All these four residues are located within the N-terminal unstructured region, fit with the Cdk1 core motif (SP or TP), and more importantly, they are conserved not only in four Schizosaccharomyces species but also in several other fungal species (Figure 4B and Supplementary Figure 2B). Interestingly, they were also proposed as Cdk1 phosphorylation sites based on quantitative phosphoproteomic analysis in a recent study (Swaffer et al., 2018) (Figure 4B). By using the phospho-specific antibodies against both pS28 and pT31, we could detect the phosphorylation of Ser28 and Thr31 in vitro by Cdc13 (cyclin B)-containing Cdk1 complexes purified from metaphase-arrested yeast cells, which was diminished when the kinase activity of ATP analogue-sensitive Cdk1 [cdc2-as1(F84G)] was inhibited by 1-NM-PP1 or when Slp1(S28A;T31A) was added as the substrate in in vitro kinase reactions (Figure 4E). Next, we examined whether Ser28 and Thr31 were phosphorylated in vivo using immunoblotting with phospho-specific antibodies to directly detect Slp1Cdc20, which was co-immunoprecipitated with Apc15 from nda3-arrested yeast cells. We observed phosphorylation of Ser28/Thr31 in wild type cells, but it disappeared following phosphatase treatment of wild type samples after immunoprecipitation, 1-NM-PP1 treatment of cdc2-as1 cells before being collected for immunoprecipitation, or when cells expressing Slp1(S28A;T31A) were used (Figure 4F). Thus, Slp1Cdc20 is also phosphorylated by Cdk1 at least at Ser28 and Thr31 in fission yeast mitotic cells.

We wondered whether Slp1Cdc20 phosphorylation by Pmk1 affects its stability and APC/C activation. Indeed, phospho-deficient Slp1T480A mutant showed elevated Slp1Cdc20 levels in both metaphase-arrested and asynchronously grown cells (Figure 4G and Supplementary Figure 7) and compromised SAC activation compared to wild type cells, likely due to more efficient APC/C activation (Figure 4H and Supplementary Figure 8). Furthermore, Slp1T480A mutant also reversed the inhibitory effect of pek1DD on Slp1Cdc20 stability and APC/C activation after release from SAC-mediated arrest (Figure 4G and 4H).

Ser76 was identified as a Pmk1-dependent phosphorylation residue in our mass spectrometry analysis, but it was considered as a Cdk1 targeting site in an earlier study (Swaffer et al., 2018). It is fairly possible that phosphorylation of Ser76 is promoted by both Cdk1 and Pmk1, which is conceivable as CDKs and MAPKs recognize the same core motif (SP/TP) (Pinna and Ruzzene, 1996), and crosstalk between CDK and MAPK has recently been reported for pheromone signaling in budding yeast (Repetto et al., 2018). We tested how the double phosphorylation sites (S76A;T480A) mutant affects Slp1Cdc20 levels or APC/C activity, and observed that the protein levels of Slp1S76A;T480A were further elevated and SAC activation efficiency in this mutant was further lowered compared to Slp1T480A mutant (Figure 4G and 4H). Of note, Slp1T480A and Slp1S76A;T480A mutant proteins became easily detected in interphase cells (Supplementary Figure 7), this is very different from wild type Slp1 which normally accumulates exclusively in mitosis (Yamada et al., 2000).

We further tested whether replacement of other identified phospho-serine/threonine residues with alanines affects Slp1Cdc20 protein levels and APC/C activity. We constructed strains carrying combinations of mutations of S28A, T31A and S59A with S76A, and also with confirmed Pmk1 phosphorylation-deficient T480A, and found that mutations of S28A, T31A and S59A did not affect Slp1Cdc20 protein levels, though they could also profoundly compromise SAC activation efficiency similarly to S76A;T480A mutant (Figure 4G, 4H and Supplementary Figure 7). We noticed that the presence of mutations of S28A, T31A and S59A also abolished the elevated Slp1Cdc20 protein levels conferred by S76A;T480A mutations (Figure 4G and Supplementary Figure 8). Thus, phosphorylation of S28, T31 and S59 targeted by Cdk1 most likely influences APC/C activation through distinct mechanism from Pmk1-mediated S76/T480 phosphorylation.

Taken together, our data suggested that activated Pmk1 phosphorylates Slp1Cdc20 in vivo, which leads to lowered Slp1Cdc20 levels, delayed SAC inactivation and APC/C activation as we observed in pek1DDcells. Also, Pmk1 can join Cdk1 to enhance the multi-site phosphorylation of their shared substrate Slp1Cdc20 and collaboratively dampen APC/C activity.

Pmk1-but not Cdk1-mediated Slp1Cdc20 phosphorylation promotes its ubiquitylation

Our above data support the notion that the direct binding of Slp1Cdc20 by Pmk1 mediates its phosphorylation, which possibly results its subsequent rapid turnover. It has been shown in both human and fission yeast cells that Apc15, one of the conserved components of APC/C, promotes Cdc20/Slp1Cdc20 autoubiquitylation and its turnover by APC/C (Mansfeld et al., 2011; May et al., 2017; Sewart and Hauf, 2017; Uzunova et al., 2012). The increased Slp1Cdc20 levels in pmk1Δ cells (Figure 2C) is reminiscent of the previous observations in apc15Δ mutant, which also shows elevated levels of Slp1Cdc20 (May et al., 2017; Sewart and Hauf, 2017). We reasoned that the absence of Apc15 may reverse the negative effect of pek1DDon Slp1Cdc20 levels and spindle checkpoint inactivation and APC/C activation. As expected, we indeed observed that deletion of apc15 recovered Slp1Cdc20 levels in pek1DD cells (Figure 5A), and also lowered the percentage of cells with Cdc13-GFP retained on SPBs in nda3-KM311 pek1DD cells after cold treatment (Figure 5B and Supplementary Figure 9A). These observations are consistent with previous reports showing the relative abundance between checkpoint proteins and the checkpoint target Slp1Cdc20 is an important determinant of checkpoint robustness (Heinrich et al., 2013) and apc15Δ mutant is spindle checkpoint-defective (May et al., 2017).

Pmk1-but not Cdk1-mediated Slp1Cdc20 phosphorylation promotes its ubiquitylation.

(A) Immunoblot analysis of Slp1Cdc20 abundance in apc15+ or apc15Δ background cells with indicated genotypes after being treated at 18 °C for 6 hr. Slp1Cdc20 levels were quantified as in Figure 2B. The experiment was repeated 3 times and the mean value for each sample was calculated. ***, p<0.001.

(B) Time-course analyses of SAC activation and inactivation in nda3-KM311 cdc13-GFP strains with indicated genotypes. The experiments were performed as in Figure 1D.

(C) Immunoblot analysis of Slp1Cdc20 abundance in K472R, K479R or K472R/K479R mutants. The experiment was repeated 3 times and the mean value for each sample was calculated. *, p<0.05; **, p<0.01; ***, p<0.001; ****, p<0.0001.

(D) Time-course analyses of SAC activation and inactivation in nda3-KM311 cdc13-GFP strains with K472R, K479R or K472R/K479R mutations. The experiments were performed as in (B).

(E) Schematic depiction of affinity pull-down assays using TUBE (Tandem Ubiquitin Binding Entity) agarose beads to detect Slp1Cdc20 ubiquitylation.

(F) TUBE pull-down assays in mts3-1 strains carrying sfGFP-tagged wild type or mutants of Slp1Cdc20 or pmk1Δ or pek1DDmutations. The TUBE beads-bound samples were blotted with anti-GFP antibodies. Asterisk indicate unspecific bands recognized by anti-GFP antibodies.

(G) Schematic summarizing the negative effect of the coupling of Pmk1 phosphorylation and K472/K479-mediated ubiquitylation on Slp1Cdc20 abundance and APC/C activation.

During the course of our mass spectrometry analysis on Slp1Cdc20 phosphorylation, we also detected Lys 479 (K479) as an ubiquitin-modifying site (Figure 4B and Supplementary Figure 5 and 6). We wondered whether K479 is involved in Slp1Cdc20 autoubiquitylation in fission yeast. In agreement with the mass spectrometry analysis, mutation of K479 to arginine indeed conferred stability of Slp1Cdc20 (Figure 5C), even in asynchronously growing cells (Supplementary Figure 10), which is reminiscent of the stabilized Slp1Cdc20 in apc15Δ mutant cells at interphase (Sewart and Hauf, 2017). These results suggested that K479 is likely Slp1Cdc20 ubiquitylation-relevant. Furthermore, nda3-KM311 slp1K479R cells also demonstrated lowered metaphase arrest rate after cold treatment (Figure 5D and Supplementary Figure 9B), indicating defective SAC activation. Previous studies have identified two neighbouring lysine residues close to the carboxy terminus of human Cdc20, Lys485 and Lys490, as ubiquitylation sites in prometaphase cells, and mutating these two lysines to arginine completely prevented Cdc20 ubiquitylation (Danielsen et al., 2011; Mansfeld et al., 2011). Very interestingly, a second lysine residue, K472, is also present in fission yeast Slp1Cdc20 adjacent to K479 (Figure 4B and Supplementary Figure 11). Although K472 was not identified as an ubiquitin-modifying residue in our mass spectrometry analysis, replacement of K472 with arginine also efficiently stabilized Slp1Cdc20, similar to K479R mutant (Figure 5C and Supplementary Figure 10). More strikingly, simultaneously mutating both K472 and K479 caused additive effects in elevating Slp1Cdc20 levels and lowering SAC activation rate in both wild type and pek1DD strain backgrounds (Figure 5C, 5D and Supplementary Figure 10), suggesting that likely both these two lysine residues at the C-terminal tail of Slp1Cdc20 contribute to its turnover via ubiquitylation, which is very similar to the mechanism in human Cdc20. Consistently, using TUBEs (tandem ubiquitin-binding entities) (Hjerpe et al., 2009) to enrich for ubiquitinated proteins in metaphase-arrested mts3-1 cells expressing sfGFP-Slp1, we were able to demonstrate that mutating both K472 and K479 largely removed higher molecular weight bands detected by immunoblotting, which were corresponding to ubiquitylated species of Slp1Cdc20 (Figure 5E, F). We also noticed that Slp1K472R; K479R rendered more severely compromised viability than either single mutants (Supplementary Figure 12), indicating blocking Slp1Cdc20 ubiquitylation could bring about detrimental consequence to the cells.

Our observation that Slp1Cdc20 levels were lowered in pek1DDmutant (Figure 2B) suggested that attenuation of Slp1Cdc20 levels upon constitutive activation of CIP signaling is very likely through facilitating its autoubiquitylation. To test this possibility, we examined and compared the ubiquitylation levels of sfGFP-Slp1 in wild type, pmk1Δ and pek1DDcells by affinity pull-down assays using TUBEs (Figure 5E). Indeed, we observed a much weaker or stronger discrete laddering pattern of sfGFP-Slp1 in pmk1Δ and pek1DD cells than that of wild type cells, indicating abated or enhanced ubiquitylation respectively (Figure 5F). Furthermore, mutations of T480A, S76A/T480A and Pmk1-docing site, but not S28A/T31A- or S28A/T31A/S59A-containing mutants, also reduced polyubiquitylation of Slp1Cdc20 (Figure 5F). All these results confirmed that Pmk1-but not Cdk1-mediated Slp1Cdc20 phosphorylation promotes its ubiquitylation.

Together, these findings lent further support to the idea that the CIP signaling pathway restrain APC/C activity through a cascade mechanism, in which Pmk1 directly binds and phosphorylates Slp1Cdc20 to facilitate its autoubiquitylation, and eventually lowers its levels (Figure 5G).

Osmotic stress and cell wall damage trigger rapid Slp1Cdc20 downregulation and mitotic exit delay

It has been well established that two MAPKs Sty1 and Pmk1 can be activated by environmental perturbations including osmotic stress, cell wall damage, thermal stress, oxidative stress, and glucose limitation, among others (Cansado et al., 2021; Madrid et al., 2006; Shiozaki and Russell, 1995b). We wondered whether the phosphorylation of Slp1Cdc20 by Pmk1 could tune the strength of SAC activation and APC/C activity in response to extracellular stress. To investigate this possibility, we performed arrest-and-release experiments with nda3-KM311 mutants and added 0.6 M KCl or 2 µg/mL of caspofungin 1 hr before release from metaphase arrest to elicit strong osmotic saline stress (KCl) or cell wall damage stress (caspofungin) response (Madrid et al., 2006; Madrid et al., 2016), respectively (Figure 6A). As previously reported, treatment with KCl or caspofungin resulted in full activation of both Pmk1 and Sty1 (for KCl treatment) or only Pmk1 (for caspofungin treatment) as determined using anti-phospho p42/44 and anti-phospho p38 antibodies, which serve as indicative of respective Pmk1 and Sty1 activation (Figure 6B). Intriguingly, treatment with KCl or caspofungin also correlated with a marked and rapid decrease in Slp1Cdc20 levels (Figure 6B).

Osmotic stress and cell wall damage trigger rapid Pmk1 phosphorylation-dependent Slp1Cdc20 downregulation and mitotic exit delay.

(A) Schematic depiction of the experimental design for treatment with KCl or caspofungin during nda3-mediated spindle checkpoint activation to activate MAPKs. Samples were collected at indicated time points for subsequent analyses including immunoblotting, co-IP and time-course analysis on SAC or APC/C activation.

(B) Immunoblot analysis of activation of MAPKs and Slp1Cdc20 protein levels. Samples with or without indicated treatments were blotted with anti-phospho p42/44 and anti-phospho p38 antibodies as indicative of phosphorylated Pmk1 (Pmk1-P) or phosphorylated Sty1 (Sty1-P), respectively. Slp1Cdc20 levels were detected with anti-Slp1 antibodies and anti-Cdc2 was used as loading control.

(C) Co-immunoprecipitation analysis of APC/C-MCC association upon environmental stress. Lid1-TAP was immunoprecipitated from nda3-KM311-arrested cells and associated Mad2-GFP, Mad3-GFP and Slp1Cdc20 were detected as in Figure 2A. Note that APC/C-MCC association was disrupted when 0.6 M KCl was present during cell culturing or during immunoprecipitation procedures.

(D) Time-course analyses of SAC activation and inactivation in nda3-KM311 cdc13-GFP strains with indicated genotypes after arrest at 18 °C and KCl or caspofungin treatments.

(E, F) Immunoblot analyses of Slp1Cdc20 abundance and time-course analyses of SAC activation and inactivation efficiency in phosphorylation- and ubiquitylation-deficient mutants upon SAC activation and environmental stress.

Surprisingly, we did not see any alteration of APC/C-MCC interaction accompanying Pmk1 activation when cells were treated by caspofungin (Figure 6C), which was different from what we observed in Padh11-pek1DD cells (Figure 2A). Also, we failed to detect enhanced APC/C-MCC association like we observed in Padh11-wis1DDcells when cells were treated by KCl, where Sty1 was activated (Figure 6C). The latter observation was most likely due to the disruptive effect of high concentration of KCl on protein-protein interactions, because we observed similar effect when we immunoprecipitated APC/C in the presence of 0.6 M KCl in the lysis buffer (Figure 6C, right panel). With this data, we could not exclude the possibility that activated Sty1 transiently enhanced APC/C-MCC association in yeast cells, which was soon counteracted by the disruptive effect of saline. Nevertheless, the high SAC activation efficiency in mitotis-arrested cells exposed to osmotic stress (KCl) or cell wall damage stress (caspofungin) was largely lowered by pmk1 deletion (Figure 6D), indicating that at least activated Pmk1 exerts regulation of the spindle checkpoint and APC/C activation under environmental stimuli. Further, Pmk1-docking site mutant slp1(DS-5A) and Pmk1 phosphorylation-deficient mutants slp1T480A and slp1S76A;T480A could partially suppress the sustained SAC activation responding to osmotic and cell wall damage stress (Figure 6E and Supplementary Figure 13).

Discussion

Involvement of MAPKs in cell cycle regulation

So far, the only known cell-cycle control stages linked to MAPKs in fission yeast are at G2/M transition and during cytokinesis (Gomez-Gil et al., 2020; Petersen and Hagan, 2005; Petersen and Nurse, 2007). The p38 MAPK family member Sty1 either arrests or promotes mitotic commitment in unperturbed, stressed or resumed cell cycles after stress depending on the downstream kinases it associates with. Sty1 can phosphorylate kinase Srk1 at threonine 463 (T463), which further phosphorylates and inhibits Cdc25 phosphatase to negatively regulate the onset of mitosis (Lopez-Aviles et al., 2005; Lopez-Aviles et al., 2008; Smith et al., 2002). Sty1 also mediates phosphorylation of Polo kinase Plo1 at serine 402 (S402), which then promotes its recruitment to the spindle pole bodies (SPBs), where it positively modulates Cdc25 control of Cdc2 (fission yeast Cdk1) activity for mitotic entry (Petersen and Hagan, 2005; Petersen and Nurse, 2007). During cytokinesis, two fission yeast formins For3 and Cdc12 cooperate to nucleate actin filaments for assembly of contractile actomyosin ring (CAR) (Coffman et al., 2013). It has been shown that Sty1 activity negatively regulates protein levels of For3 in response to actin cytoskeleton damage to downregulate CAR assembly (Gomez-Gil et al., 2020). Very recently, Pmk1 has also been proposed to act as a negative regulator of cytokinesis, which delays CAR constriction upon cell wall stress (e.g. treatment with blankophor, caspofungin and caffeine) through a novel checkpoint mechanism (Edreira et al., 2020). However, the detailed mechanisms how Sty1 and Pmk1 pose negative effect on cytokinesis have not been elucidated.

In the current study, we have expanded MAPK signaling-controlled cell-cycle stage to metaphase-anaphase transition in fission yeast. In response to unattached or tensionless kinetochores, the SAC generates the MCC, which inhibits the APC/C and block anaphase onset by suppressing APC/C-catalyzed ubiquitylation and subsequent degradation of securin and cyclin B. As a critical APC/C inhibitor, MCC fulfills its function through the physical binding of BubR1/Mad3 and Mad2 to two molecules of Cdc20/Slp1Cdc20 (i.e. Cdc20MCC and Cdc20APC/C), and this mechanism has been confirmed at least in both humans and fission yeast (Alfieri et al., 2016; Chao et al., 2012; Izawa and Pines, 2015; Sewart and Hauf, 2017; Yamaguchi et al., 2016). In this study, we have uncovered a dual mechanism, in which two fission yeast MAPKs Pmk1 and Sty1 restrain APC/C activity through distinct manners, either by phosphorylating Slp1Cdc20 to lower its levels or by phosphorylating unknown substrate(s) to promote MCC-APC/C association and thus SAC signaling strength, respectively (Figure 7). Therefore, our study has established another critical APC/C-inhibitory mechanism in the spindle checkpoint, though it will be important for future studies to identify the phosphorylation target(s) posed by SAP signaling pathway. Nevertheless, our current study together with earlier report (Edreira et al., 2020) have implicated fission yeast Pmk1 in both the mitotic and cytokinesis surveillance mechanisms in response to unfavorable environmental stimuli.

Summary of the mechanisms that how MAPKs negatively regulate APC/C activity in fission yeast.

(A) Schematic depiction of a possible dual mechanism that how activated MAPK signaling pathways are involved in delaying APC/C activation and SAC inactivation. Upon SAC and MAPK signaling activation, division of labor between the CIP and the SAP pathways enables the phosphorylation of Slp1Cdc20 and unidentified substrate(s) by Pmk1 and Sty1 respectively, which leads to lowered Slp1Cdc20 level and enhanced MCC affinity for APC/C.

(B) Summary of the alteration of Slp1Cdc20 protein levels and MCC-APC/C association strength under various conditions examined in this study. White wavy line indicates the absence of Apc15.

The regulation of mitotic entry by MAPK in fission yeast is triggered by many environmental insults, including high osmolarity, oxidative stress, heat shock, centrifugation, nutrient starvation and rapamycin treatment (Hartmuth and Petersen, 2009; Petersen and Hagan, 2005; Petersen and Nurse, 2007; Shiozaki and Russell, 1995a; Shiozaki et al., 1998). In this study, we found that treatment of fission yeast cells with KCl or caspofungin, which provokes a strong osmotic stress or cell wall damage response and quick Pmk1 activation, correlated with an attenuated Slp1Cdc20 protein levels and delayed APC/C activation. Deletion of pmk1+ largely overrides the negative effects of KCl or caspofungin treatment on APC/C activation (Figure 6D), confirming that activated Pmk1 is responsible for prolonged metaphase arrest under environmental stress. Thus, our finding implicates MAPK Pmk1 as an important regulator for anaphase entry in response to adverse extracellular conditions. Intriguingly, one previous study showed that treatment of human cells with the carcinogen cadmium activates MAPK p38 (fission yeast Sty1 homologue) and triggers accelerated ubiquitination and proteolysis of Cdc20, which is essential for prometaphase arrest under SAC (Yen and Yang, 2010). Therefore, although different MAPKs might be employed to fulfil the task in yeast and higher eukaryotes, it is still possible that phosphorylation and regulation of Cdc20 levels by MAPKs is an evolutionarily conserved mechanism to adjust mitotic exit to extracellular or environmental stimuli.

Negative phospho-regulation of Cdc20 by MAPK as a spindle assembly checkpoint-dependent APC/C-inhibitory mechanism

APC/C activity can be regulated at multiple levels, Cdk1-dependent APC/C phosphorylation is one of the major prerequisites for APC/C activation and anaphase onset, in which Apc3 and Apc1 are key subunits for positive phospho-regulation (Fujimitsu et al., 2016; Kraft et al., 2003; Qiao et al., 2016; Steen et al., 2008; Zhang et al., 2016). However, it has been shown that Cdk1-dependent phosphorylation of APC/C co-activator Cdc20 lessens its interaction with APC/C, thus serves as a negative phospho-regulation (Labit et al., 2012). In addition to its phosphorylation by Cdk1, phosphorylation of Cdc20 by multiple other kinases, such as Bub1, PKA, MAPK and Plk1, has also been demonstrated or suggested in Xenopus, humans or budding yeast to be inhibitory to APC/C activation and improper anaphase onset upon SAC activation ((Jia et al., 2016) and reviewed in (Yu, 2007)). Therefore, Cdc20 generally serves as an integrator of multiple intracellular signaling cascades that regulate progression through mitosis. In this work, we demonstrated that phosphorylation of Slp1Cdc20 by Pmk1 but not Cdk1 promotes its autoubiquitylation, this mechanism is distinct from that reported in Xenopus egg extracts and tadpole cells, in which phosphorylation of Cdc20 by MAPK is instead responsible for facilitated Cdc20 binding by MCC to prevent it from activating the APC/C (Chung and Chen, 2003).

Although multiple kinases have been proposed to be directly involved in phosphorylation of Cdc20 proteins (Yu, 2007), precisely identifying the in vivo phosphorylation sites in Cdc20 has turned out to be challenging, most likely due to its protein scarcity and instability. Based on known Cdk1 core consensus (SP/TP), 9 residues in human Cdc20 were originally suggested as putative Cdk1 sites, but among them only Ser41 has been proven to be phosphorylated in vivo by mass spectrometry analysis (Kramer et al., 2000; Yudkovsky et al., 2000). However, subsequent mass spectrometry analyses identified Ser41 as one of the Bub1 phosphorylation sites both in vivo and in vitro and another putative Cdk1 site Thr106 as one of the Plk1 sites (Jia et al., 2016; Tang et al., 2004). It is very likely that there are not as many genuine Cdk1 sites in Cdc20 as originally assumed. One supporting evidence is that mutation of only one (Thr32) of the three putative Cdk sites to alanine in N terminus of C. elegans CDC-20 is sufficient to phenocopy 3A mutant for accelerated APC/C activation (Kim et al., 2017). Interestingly, Thr32 of C. elegans CDC-20 is conserved as Thr 59 and Thr68 in human and Xenopus Cdc20, respectively (see Supplementary Figure 14). Based on two-dimensional tryptic phosphopeptide analysis, Thr68 in Xenopus Cdc20 has also been regarded as a MAPK target site (Chung and Chen, 2003). These studies in diverse organisms raised an issue that whether multiple kinases share the same phosphorylation sites in Cdc20? It would not be so surprising if it turns out to be the case eventually, at least for Cdk1 and MAPKs, because CDKs and MAPKs recognize the same core motif (SP/TP) (Pinna and Ruzzene, 1996).

By purifying Apc15-aasociated fission yeast Slp1Cdc20 proteins from checkpoint-arrested pek1DDand pmk1Δ cells, we were able to identify Thr480 and Ser76 as in vivo Pmk1-relevant phosphorylation sites and 3 Pmk1-independent phosphorylation sites (S28, T31, S59) based on mass spectrometry analyses. All our identified sites except Thr480 are located in the N terminal disordered portion of Slp1Cdc20 sequence, this is similar to the phosphorylation residues identified by biochemical and functional analyses in other organisms, in which the major phosphorylation sites all fall within the N-terminus of Cdc20 homologues. We further provided evidence that S28/T31 are phosphorylated by Cdk1 both in vitro and in vivo, and T480 phosphorylated by Pmk1 both in vitro and in vivo. We noticed that there is a tendency for reported phosphorylation residues to be concentrated in the N terminus of Cdc20 proteins (Jia et al., 2016; Tang et al., 2004; Yu, 2007). However, due to the lack of clear sequence homology in their N-terminal and C-terminal tails between Slp1Cdc20 and its Cdc20 homologues in higher eukaryotes (see Supplementary Figure 14), it is impossible to certainly assign any identified sites from different organisms as homologous sites. But it is possible that multiple kinases modify shared sites in Cdc20, this collaborative effect may be especially helpful under adverse environmental stress. Nevertheless, all reported Cdc20 phosphorylation events constitute an APC/C-inhibitory mechanism, no matter what kinases are responsible for the modifications (Chung and Chen, 2003; D’Angiolella et al., 2003; Jia et al., 2016; Kim et al., 2017; Kramer et al., 2000; Labit et al., 2012; Tang et al., 2004; Yu, 2007; Yudkovsky et al., 2000).

The current model of inhibitory signaling from Cdc20 posits that its phosphorylation either promotes the formation of MCC or inhibits APC/CCdc20 catalytically, both are required for proper spindle checkpoint signaling (Jia et al., 2016; Labit et al., 2012; Yamano, 2019). Interestingly, our current study suggests Slp1Cdc20 phosphorylation by fission yeast MAPK Pmk1 may represent another critical APC/C-inhibitory mechanism in the spindle checkpoint through reducing the protein levels of Slp1Cdc20, which is achieved via autoubiquitylation-mediated degradation (Figure 7). Actually, previous studies in fission and budding yeast have underlined the importance of accurate relative abundance between checkpoint proteins and Cdc20/Slp1Cdc20, which sets an important determinant of checkpoint robustness (Heinrich et al., 2013; Pan and Chen, 2004). A benefit of allowing the association of MCC-bound Cdc20 with APC/C is to provide an opportunity for Cdc20 autoubiquitylation by APC/C, a process aided by APC/C subunit Apc15 (Mansfeld et al., 2011; May et al., 2017; Sewart and Hauf, 2017; Uzunova et al., 2012). Notably, the depletion of apc15+ restores Slp1Cdc20 levels in pek1DD cells (Figure 5A), consistent with the idea that Pmk1-mediated Slp1Cdc20 phosphorylation very likely facilitates its autoubiquitylation. Indeed, our biochemical and functional analyses further establish K479 and K472 in Slp1Cdc20 as two key residues directly involved in its ubiquitylation, which likely follows its phosphorylation, particularly at adjacent Thr480. Although K485 and K490 are similarly located at C terminus of human Cdc20 and can be ubiquitylated and responsible for Cdc20 degradation (Danielsen et al., 2011; Mansfeld et al., 2011), there has been no reports on phosphorylation-modified residues in this region. However, although the sequence homology within N- and C-terminal regions between Slp1Cdc20 and Cdc20 homologues in higher eukaryotes is extremely low, we do notice that there are several lysine-flanking serine or threonine residues within Cdc20 C-terminal tails of higher eukaryotes (see Supplementary Figure 11), this raises a possibility that this region may be involved in phosphorylation-coupled ubiquitylation. Currently, we do not know how T480 phosphorylation accelerate ubiquitylation of K479 and K472 in Slp1Cdc20. It is possible that phosphorylation of Slp1Cdc20 might alter the mode of Cdc20 binding to APC/C which becomes conducive to catalyze Slp1Cdc20 ubiquitylation.

In summary, our studies revealed a distinct dual mechanism of MAPK-dependent anaphase onset delay imposed directly on APC/C co-activator Slp1Cdc20 and an unidentified substrate, which may emerge as an important regulatory layer to fine tune APC/C and MCC activity upon SAC activation. Importantly, although this mechanism of APC/C regulation by two MAPKs has not been examined in other organisms, it is very likely widespread across eukaryotes including humans, given the involvement of MAPKs in general response to wide range of adverse environmental situations and similar regulations of APC/C and SAC in eukaryotes characterized so far. It is also plausible to assume that the MAPK-mediated APC/C inhibition pathway may also contribute to the pathogenesis of multiple cancers. Indeed, Cdc20 upregulation has been demonstrated in several types of cancers, including breast and colon cancer, and it contributes to aggressive tumor progression and poor prognosis in gastric cancer and primary non-small cell lung cancer (Chi et al., 2019). It would be interesting to address whether the altered levels of Cdc20 in these cancers are MAPK relevant. Nevertheless, inhibition of APC/CCdc20 activity or induction of Cdc20 degradation have already been listed as alternative therapeutic strategies to control cancer (Greil et al., 2022; Jeong et al., 2022). Thus, although the detailed molecular mechanisms may vary, further studies on MAPK-mediated APC/CCdc20 inhibition mechanism in human would lead to better understanding of its oncogenic role in tumor progression.

Materials and Methods

Fission yeast strains, media and genetic methods

Fission yeast cells were grown in either YE (yeast extract) rich medium or EMM (Edinburgh minimal medium) containing the necessary supplements. G418 disulfate (Sigma-Aldrich; A1720), hygromycin B (Sangon Biotech; A600230) or nourseothiricin (clonNAT; Werner BioAgents; CAS#96736-11-7) was used at a final concentration of 100 μg/ml and thiabendazole (TBZ) (Sigma-Aldrich; T8904) at 5-15 μg/ml in YE media where appropriate. For serial dilution spot assays, 10-fold dilutions of a mid-log-phase culture were plated on the indicated media and grown for 3 to 5 days at indicated temperatures.

To create strain with an ectopic copy of slp1+ at lys1+ locus on chromosome 1, the open reading frame of slp1+(1467 bp) and its upstream 1504 bp sequence was first cloned into the vector pUC119-Padh21-MCS-hphMX6-lys1* with adh21 promoter (Padh21) removed using the ‘T-type’ enzyme-free cloning method (Chen et al., 2017). The sequence of hphMX6 in above plasmid was replaced by fragment corresponding to nourseothiricin resistance gene sequence by BglII-EcoRI cut and re-ligation, generating pUC119-Pslp1-slp1+-Tadh1::natMX6-lys1*. The resultant plasmids were linearized by ApaI and integrated into the lys1+locus, generating the strains lys1Δ::Pslp1-slp1+-Tadh1::hphMX6 and lys1Δ::Pslp1-slp1+-Tadh1::natR.

To generate the mutant strains harboring lysine/arginine (K/R) or serine/threonine to glutamic acid (E) or alanine (A) mutations, or lysine (K) to arginine (R) mutations within Slp1, the desired mutations were introduced into pUC119-Pslp1-slp1+-Tadh1::hphMX6 using standard methods of site-directed mutagenesis.

To generate the strains carrying superfolder GFP (sfGFP)-tagged wild type or mutant Slp1, a fragment corresponding to the sequence of sfGFP was inserted in front of slp1 in pUC119-Pslp1-slp1-Tadh1::hphMX6-lys1* using the ‘T-type’ enzyme-free cloning method. The resulting plasmids were linearized and integrated into the lys1+ locus of chromosome 1 using the hygr marker. Then endogenous slp1+ was deleted and replaced with ura4+.

To generate the strains expressing ade6::Padh1-GST-slp1(456-488aa)::natR, C-terminal sequence of slp1+corresponding to 456-488aa was first cloned into the vector pUC119-Padh1-MCS-natMX6-lys1* and the sequence of lys1* was replaced with ade6+ from pAde6-NotI-hphMX (a kind gift of Li-lin Du), and then the sequence of GST from pGEX-4T-1 was inserted in front of slp1. The resulting plasmid pUC119-Padh1-GST-slp1(456-488aa)-natMX6-ade6 was linearized and integrated into the ade6+ locus of chromosome III using the natr marker.

To generate the constitutively active MAPKK strains pek1DDor wis1DD, the Ser234 and Thr238 of pek1+ in pUC119-Padh21-6HA-pek1+-Tadh1-hphMX6-lys1*, pUC119-Padh11-6HA-pek1+-Tadh1-hphMX6-lys1* and pUC119-Padh11-6HA-pek1+-Tadh1-kanMX6-ura4+and the Ser469 and Thr473 of wis1+ in pUC119-Padh21-6HA-wis1+-Tadh1-hphMX6-lys1*, pUC119-Padh11-6HA-wis1+-Tadh1-hphMX6-lys1* and pUC119-Padh11-6HA-wis1+-Tadh1-kanMX6-ura4+were changed to aspartic acids (D) by standard methods of site-directed mutagenesis. The resulting plasmids were linearized and integrated into the lys1+ locus of chromosome I or the ura4+ locus of chromosome III using the kanrmarker. A list of the yeast strains used is in Supplementary Table S1.

Fission yeast cell synchronization methods

For nda3-KM311 strains, cells were grown at the permissive temperature for nda3-KM311 (30 °C) to mid-log phase, synchronized at S phase by adding HU (Sangon Biotech; A600528) to a final concentration of 12 mM for 2 hours followed by a second dose of HU (6 mM final concentration) for 3.5 hours. HU was then washed out and cells were released at specific temperatures as required by subsequent experiments.

Spindle checkpoint activation and silencing assays in fission yeast

For checkpoint silencing assay in the absence of microtubules, mid-log cdc13-GFP nda3-KM311 cells were first synchronized with HU at 30 °C and then arrested in early mitosis by shifting to 18 °C for 6 hours, followed by incubation at 30 °C to allow spindle reformation and therefore spindle checkpoint inactivation. For osmotic stress or cell wall damage stress experiments, 0.6 M KCl or 2 μg/mL caspofungin was added into cultures during 18 °C incubation 1 hr before being shifted to 30 °C. Cells were withdrawn at certain time intervals and fixed with cold methanol and stained with DAPI. 200-300 cells were analyzed for each time point. Each experiment was repeated at least three times.

Immunoblotting and immunoprecipitation

For routine Western blot and immunoprecipitation experiments, yeast cells were collected from nda3-KM311 arrested and released cultures, followed by lysing with glass bead disruption using Bioprep-24 homogenizer (ALLSHENG Instruments, Hangzhou, China) in NP40 lysis buffer (6 mM Na2HPO4, 4 mM NaH2PO4, 1% NP-40, 150 mM NaCl, 2 mM EDTA, 50 mM NaF, 0.1 mM Na3VO4) plus protease inhibitors as previously described (Wang et al., 2012). Proteins were immunoprecipitated by IgG Sepharose beads (GE Healthcare; 17-0969-01) (for Lid1-TAP) or anti-GFP nanobody agarose beads (AlpalifeBio, Shenzhen, China; KTSM1301) (for Mad3-GFP). Immunoblot analysis of cell lysates and immunoprecipitates was performed using appropriate antibodies at 1:500 - 1:5000 dilutions and was read out using chemiluminescence.

Detection of in vivo phosphorylation of Slp1Cdc20 at Thr480 and Ser28/Thr31

For detection of phosphorylation of Slp1Cdc20 at Thr480 in vivo, mts3-1 mutant strains expressing ade6::Padh1-GST-slp1(456-488aa)::natR in wild type, pmk1Δ and lys1Δ::Padh11-6xHA-pek1DDbackground were arrested at 36℃ for 3.5 hours, followed by pull-down of GST-Slp1(456-488aa) by glutathione Sepharose 4B (GE Healthcare). Purified GST-Slp1(456-488aa) was detected with either goat anti-GST HRP-conjugated antibodies (GE Healthcare) or rabbit polyclonal anti-pThr480 antibodies.

For detection of phosphorylation of Slp1Cdc20 at Ser28/Thr31 in vivo, nda3-KM311 apc15-13myc::kanR strains expressing slp1+ or slp1(S28A;T31A) in wild type or cdc2-asM17 background were first synchronized with HU at 30 °C and then arrested in early mitosis by shifting to 18 °C for 6 hours. For cdc2-asM17 strains, 1 μM 1-NM-PP1 (Toronto Research Chemicals; A603003) was added into one of the cultures to inactivate Cdc2 (Cdk1) 20 min before being collected. Apc15-13myc was immunoprecipitated by anti-c-Myc magnetic beads (MedChemExpress (MCE); HY-K0206). To remove Slp1Cdc20 phosphorylation, one of the immunoprecipitated samples was treated with 400 units of Lambda protein phosphatase (New England Biolabs, P0753S) at 30 °C for 30 min. Co-immunoprecipitated Slp1Cdc20 was detected with rabbit polyclonal anti-Slp1 antibodies, and the phosphorylated form of Slp1Cdc20 was detected with rabbit polyclonal anti-pSer28/pThr31 antibodies respectively.

Purification of APC/C and analyses by mass spectrometry

To prepare APC/C for mass spectrometry analyses, Apc15-GFP was purified from 6-liter cultures of nda3-KM311 arrested cells with desired genetic background. Cells were disrupted and cell lysates were prepared as described above for routine immunoprecipitation experiments.

For mass spectrometry analyses, purified samples were first run on PAGE gels, after staining of gels with Coomassie blue, excised gel segments were subjected to in-gel trypsin (Promega, V5111) digestion and dried. Samples were then analyzed on a nanoElute (plug-in V1.1.0.27; Bruker, Bremen, Germany) coupled to a timsTOF Pro (Bruker, Bremen, Germany) equipped with a CaptiveSpray source. Peptides were separated on a 15cm X 75μm analytical column, 1.6 μm C18 beads with a packed emitter tip (IonOpticks, Australia). The column temperature was maintained at 55 °C using an integrated column oven (Sonation GmbH, Germany). The column was equilibrated using 4 column volumes before loading sample in 100% buffer A (99.9% MilliQ water, 0.1% FA) (Both steps performed at 980 bar). Samples were separated at 400 nL/min using a linear gradient from 2% to 25% buffer B (99.9% ACN, 0.1% FA) over 90 min before ramping to 37% buffer B (10min), ramp to 80% buffer B (10min) and sustained for 10 min (total separation method time 120 min). The timsTOF Pro (Bruker, Bremen, Germany) was operated in PASEF mode using Compass Hystar 6.0. Mass Range 100 to 1700m/z, 1/K0 Start 0.6 V⋅s/cm2 End 1.6 V⋅s/cm2, Ramp time 110.1ms, Lock Duty Cycle to 100%, Capillary Voltage 1600V, Dry Gas 3 L/min, Dry Temp 180°C, PASEF settings: 10 MS/MS scans (total cycle time 1.27sec), charge range 0-5, active exclusion for 0.4 min, Scheduling Target intensity 10000, Intensity threshold 2500, CID collision energy 42eV. All raw files were analyzed by PEAKS Studio Xpro software (Bioinformatics Solutions Inc., Waterloo, ON, Canada). Data was searched against the S. pombe proteome sequence database (Uniprot database with 5117 entries of protein sequences at https://www.uniprot.org/proteomes/UP000002485). De novo sequencing of peptides, database search and characterizing specific PTMs were used to analyze the raw data; false discovery rate (FDR) was set to ≤1%, and [-10*log(p)] was calculated accordingly where p is the probability that an observed match is a random event. The PEAKS used the following parameters: (i) precursor ion mass tolerance, 20 ppm; (ii) fragment ion mass tolerance, 0.05 Da (the error tolerance); (iii) tryptic enzyme specificity with two missed cleavages allowed; (iv) monoisotopic precursor mass and fragment ion mass; (v) a fixed modification of cysteine carbamidomethylation; and (vi) variable modifications including N-acetylation of proteins and oxidation of Met.

Ubiquitin affinity pull-down assays

For detection of in vivo ubiquitylation of Slp1Cdc20, strains all carried mts3-1 mutation were used. sfGFP-tagged wild type Slp1 or Slp1 mutants were expressed in the genomically integrated form of Pslp1-sfGFP-slp1(WT or mutants)::hygR in wild type, pmk1Δ or Padh11-6xHA-pek1DDbackground. Cultures were first grown at 25 °C to mid-log phase and then shifted to 36℃ for 3.5 hours to block cells in mitosis prior to harvesting and cell lysis. Ubiquitinated proteins were pulled down from yeast lysates using Tandem Ubiquitin Binding Entities (TUBEs) (LifeSensors, UM402) according to the manufacturer’s instructions. The bound proteins were eluted from washed Agarose-TUBE2 beads and analyzed by western blotting using a mouse monoclonal anti-GFP antibody (Beijing Ray Antibody Biotech; RM1008). Cell extracts were also immunoblotted for GFP and Cdc2 as loading controls (inputs).

Expression and purification of recombinant proteins

GST or MBP fusion constructs of fission yeast full-length or truncated Slp1Cdc20, Pmk1, Pek1 and Atf1 were generated by PCR of the corresponding gene fragments from yeast genomic DNA or cDNA and cloned in frame into expression vectors pGEX-4T-1 (GE Healthcare) or pMAL-c2x (New England Biolabs), respectively. Site-directed mutagenesis was done by standard methods to generate vectors carrying Pek1-DD(S234D, T238D) or Slp1Cdc20 harboring K(Lys)/R(Arg) to E(Glu) or A(Ala) mutations at individual or combined sites including K19, K20, R21, K47, R48, K140, K141, K161, R162, R163, K472, R473. Integrity of cloned DNA was verified by sequencing analysis. All recombinant GST- or MBP-fusion proteins were expressed in Escherichia coli BL21 (DE3) cells. The recombinant proteins were purified by glutathione Sepharose 4B (GE Healthcare) or Amylose Resin High Flow (New England Biolabs) respectively according to the manufacturers’ instructions.

In vitro pulldown assay

The recombinant MBP-Slp1Cdc20 and MBP, or GST-Pmk1 and GST proteins were produced in bacteria and purified as described above. To detect the interaction between bacterially expressed Slp1Cdc20 and Pmk1-HA expressed in yeast, about 1 μg of MBP-Slp1Cdc20 or MBP immobilized on amylose resin was incubated with cleared cell lysates prepared from yeast strains nda3-KM311 pmk1-6His-HA, which were HU synchronized and arrested at prometaphase by being grown at 18 °C for 6 hours. To detect the direct interaction between bacterially expressed Slp1 and Pmk1, about 1 μg of GST-Pmk1 or GST immobilized on glutathione Sepharose 4B resin was incubated with cleared bacterial lysates expressing MBP-Slp1Cdc20. Samples were incubated for 1-3 hours at 4 °C. Resins were washed 3 times with lysis buffer (6 mM Na2HPO4, 4 mM NaH2PO4, 1% NP-40, 150 mM NaCl, 2 mM EDTA, 50 mM NaF, 0.1 mM Na3VO4, plus protease inhibitors), suspended in SDS sample buffer, and then subject to SDS-PAGE electrophoresis and Coomassie brilliant blue (CBB) staining.

Non-radioactive in vitro kinase assay

GST-Slp1Cdc20, GST-Atf1, GST-Pmk1, and GST-Pek1DD fusions were expressed and purified from Escherichia coli with glutathione Sepharose 4B as described above. Pmk1-HA-His and Cdc13-GFP were purified from yeast lysates by immuprecipitation with anti-HA or anti-GFP antibodies respectively.

For detection of phosphorylation of Slp1Cdc20 and Atf1 by yeast derived Pmk1, glutathione Sepharose 4B beads bound with GST-Slp1Cdc20 or GST-Atf1 and agarose beads bound with Pmk1-HA-His were mixed and washed 3 times with kinase buffer (50 mM Tris-HCl, pH 7.5, 10 mM MgCl2, 1 mM EGTA), then the reactions were incubated in kinase buffer with 20 μM ATPγS (Sigma-Aldrich; A1388) at 30°C for 45 min. The kinase reactions were stopped by adding 20 mM EDTA, and the reaction mixture was alkylated after incubation at room temperature with 2.5 mM p-nitrobenzyl mesylate (PNBM) (Abcam; ab138910) for 1 hr. The alkylation reaction was stopped by boiling in SDS-PAGE loading buffer. Phosphorylated proteins were detected with rabbit monoclonal anti-Thiophosphate ester antibody (Abcam, ab239919).

For detection of phosphorylation of Slp1Cdc20 at Thr480 by bacterially expressed Pmk1, beads bound with GST-Pmk1, GST-Pek1(DD) and fused substrates GST-Slp1(456-488aa) or GST-Slp1(456-488aa, T480A) were washed and incubated in kinase buffer (50 mM Tris-HCl, pH 7.5, 10 mM MgCl2, 1 mM EGTA) with 10 mM ATP (Sangon Biotech; A600020-0005) at 30°C for 45 min. The reaction was stopped by boiling in SDS-PAGE loading buffer. Phosphorylated proteins were detected with rabbit polyclonal anti-pThr480 antibodies.

For detection of phosphorylation of Slp1Cdc20 at Ser28/Thr31 by Cdk1, beads bound with bacterially expressed MBP-Slp1(1-190aa) or MBP-Slp1(1-190aa,S28A;T31A) and agarose beads bound with Cdc13-GFP/Cdc2 from yeast lysate were mixed and washed 3 times with kinase buffer (10 mM Tris-HCl, pH 7.4, 10 mM MgCl2, 1 mM DTT), then the reactions were incubated in kinase buffer with 10 mM ATP (Sangon Biotech; A600020-0005) at 30°C for 45 min. To inactivate Cdc2-as1, 5 μM 1-NM-PP1 (Toronto Research Chemicals; A603003) was added into the reaction. The reaction was stopped by boiling in SDS-PAGE loading buffer. Phosphorylated proteins were detected with rabbit polyclonal anti-pSer28/pThr31 antibodies.

Antibodies for immunoblotting

The phospho-site specific antibodies were made in an in-house facility (for anti-Slp1-pS28/pT31) or by ABclonal Technology Co., Ltd. (Wuhan, China) (for anti-Slp1-pT480) by immunizing rabbits with pS28/pT31- or pT480-containing peptides (LVFPN(S-p)PI(T-p)PLHQQ or PITK(T-p)PSSS, respectively) coupled to haemocyanin (Sigma). The purified antibodies were concentrated to 0.2 mg/mL and used at 1:100 dilution. The following antibodies used for immunoblot analyses were purchased from the indicated commercial sources and were used at the indicated dilution: peroxidase-anti-peroxidase (PAP) soluble complex (Sigma-Aldrich; P1291; 1:10,000); rabbit polyclonal anti-Myc (GeneScript; A00172-40; 1:2,000); mouse monoclonal anti-GFP (Beijing Ray Antibody Biotech; RM1008; 1:2,000); rat monoclonal anti-HA (Roche, Cat. No. 11 867 423 001; 1:2,000); rabbit polyclonal anti-Slp1 (Kim et al., 1998) (1:500); rabbit polyclonal anti-phospho-p44/42 (detecting activated Pmk1) (Cell Signaling Technology; #9101; 1:1,000); rabbit monoclonal anti-phospho-p38 (Cell Signaling Technology; #4511; 1:1,000); rabbit polyclonal anti-PSTAIRE (detecting Cdc2) (Santa Cruz Biotechnology; sc-53; 1:1,000); rabbit monoclonal anti-Thiophosphate ester antibody (Abcam; ab239919; 1:5,000); goat anti-GST HRP-conjugated antibody (RRID:AB_771429; GE Healthcare; 1:10,000). Secondary antibodies used were goat anti-mouse or goat anti-rabbit polyclonal IgG (H+L) HRP conjugates (Thermo Fisher Scientific; #31430 or #32460; 1:10,000).

RT-qPCR

For RT-qPCR of slp1+ mRNA, nda3-KM311 strains were grown, synchronized by HU and arrested at metaphase with SAC being activated as described above. Cells were collected and total cellular RNA was isolated by using TRIZOL method. cDNA was prepared using oligo(dT) and PrimeScript reverse transcriptase (Takara, D2680A). Real-time PCR was performed in the presence of SYBR Green using an Applied Biosystems 7900HT light cycler. Relative RNA levels were calculated using the ΔCT method and normalized to act1+ levels.

Fluorescence microscopy

Cdc13-GFP proteins were observed in cells after fixation with cold methanol. For DAPI staining of nuclei, cells were fixed with cold methanol, washed in PBS and resuspended in PBS plus 1 μg/ml DAPI. Photomicrographs of cells were obtained using a Nikon 80i fluorescence microscope coupled to a cooled CCD camera (Hamamatsu, ORCA-ER) or a Perkin Elmer spinning-disk confocal microscope (UltraVIEW® VoX) with a 100x NA 1.49 TIRF oil immersion objective (Nikon) coupled to a cooled CCD camera (9100-50 EMCCD; Hamamatsu Photonics) and spinning disk head (CSU-X1, Yokogawa). Image processing and analysis were carried out using Element software (Nikon), ImageJ software (National Institutes of Health) and Adobe Photoshop.

Statistical analyses and reproducibility

No statistical methods were used to predetermine sample size. Sample size was as large as practicable for all experiments. The experiments were randomized and the investigators were blinded to group allocation during experiments but not blinded to outcome assessment and data analysis.

All experiments were independently repeated two to more than three times with similar results obtained. For quantitative analyses of each time course experiment, 300 cells were counted for each time point or sample. The same sample was not measured or counted repeatedly. No data were excluded from our studies. Data collection and statistical analyses were performed using Microsoft Office Excel or GraphPad Prism 6 softwares. Data are expressed as mean values with error bars (± standard deviation/s.d.) and were compared using two-tailed Student’s t-tests unless indicated otherwise.

eLife Assessment

The mechanisms that ensure accurate chromosome segregation are key for genome integrity and defects therein can cause human disease. Although the involvement of MAP kinases in modulating mitosis is known, this manuscript makes a valuable contribution by going to some lengths to reveal links between Spindle Assembly Checkpoint dynamics and stress-responsive MAP-kinase pathways. The strength of the evidence is solid but there are minor weaknesses, which need to be addressed.

Reviewer #1 (Public Review):

Summary:

This manuscript addresses two main issues:

(i) do MAPKs play an important role in SAC regulation in single-cell organism such as S pombe?

(ii) what is the nature of their involvement and what are their molecular targets?

The authors have extensively used the cold-sensitive β-tubulin mutant to activate or inactivate SAC employing an arrest-release protocol. Localization of Cdc13 (cyclin B) to the SPBs is used as a readout for the SAC activation or inactivation. The roles of two major MAPK pathways i.e. stress-activated pathway (SAP) and cell integrity pathway (CIP), have been explored in this context (with CIP more extensively than SAP). Sty1Δ or pmk1Δ mutants were used to inactivate the SAP or CIP pathways and wis1DD or pek1DD expression was utilized to constitutively activate these pathways, respectively. Lowering of Slp1Cdc20 abundance (by phosphorylation of Slp1-Thr 480) is revealed as the main function of MAPK to augment the robustness of the spindle assembly checkpoint.

Strengths:

The experiments are generally well-conducted, and the results support the interpretations in various sections. The experimental data clearly supports some of the key conclusions:

(1) While inactivation of SAP and CIP compromises SAC-imposed arrest, their constitutive activation delays the release from the SAC-imposed arrest.

(2) CIP signaling, but not SAP signaling, attenuates Slp1Cdc20 levels.

(3) Pmk1 and Cdc20 physically interact and Pmk1-docking sequences in Slp1 (PDSS) are identified and confirmed by mutational/substitution experiments.

(4) Thr480 (and also S76) is identified as the residue phosphorylated by Pmk1. S28 and T31 are identified as Cdk1 phosphorylation sites. These are confirmed by mutational and other related analyses.

(5) Functional aspects of the phosphorylation sites have been elucidated to some extent: (a) Phosphorylation of Slp1-T480 by Pmk1 reduces its abundance thereby augmenting the SAC-induced arrest (b) S28, T31 (also S59) are phosphorylated by Cdk1(c) K472 and K479 residues are involved in ubiquitylation of Slp.

Weaknesses:

(1) Cdc13 localization to SPBs has been used as a readout for SAC activation/inactivation throughout the manuscript. However, the only image showing such localization (Figure 1C) is of poor quality where the Cdc13 localization to SPBs is barely visible. This should be replaced by a better image.

(2) The overlapping error bars in Cdc13-localization data in some figures (for instance Figure 3E and 4H) make the effect of various mutations on SAC activation/inactivation rather marginal. In some of these cases, Western-blotting data support the authors' conclusions better.

(3) This specific point is not really a weakness but rather a loose end:

One of the conclusions of this study is that MAPK (PMK1) contributes to the robustness of SAC-induced arrest by lowering the abundance of Slp1Cdc20. The authors have used pmk1Δ or constitutively activating the MAPK pathways (Pek1DD) and documented their effect on SAC activation/inactivation dynamics. It is not clear if SAC activation also leads to activation of MAPK pathways for them to contribute to the SAC robustness. To tie this loose end, the author could have checked if the MAPK pathway is also activated under the conditions when SAC is activated. Unless this is shown, one must assume that the authors are attributing the effect they observe to the basal activity of MAPKs.

(4) This is also a loose end:

The authors show that activation of stress pathways (by addition of KCl for instance) causes phosphorylation-dependent Slp1Cdc20 downregulation (Figure 6) under the SAC-activating condition. Does activation of the stress pathway cause phosphorylation-dependent Slp1Cdc20 downregulation under the non-SAC-activation condition or does it occur only under the SAC-activating condition?

(5) Although the authors have gone to some length to identify S28 and T31 (also S59) as phosphorylation sites for Cdk1, their functional significance in the context of MAPK involvement is not yet clear. Perhaps it is outside the scope of this study to dig deeper into this aspect more than the authors have.

(6) In its current state, the Discussion section is quite disjointed. The first section "Involvement of MAPKs in cell cycle regulation" should be in the Introduction section (very briefly, if at all). It certainly does not belong to the Discussion section. In any case, the Discussion section should be more organized with a better flow of arguments/interpretations.

Reviewer #2 (Public Review):

Summary:

This study by Sun et al. presents a role for the S. pombe MAP kinase Pmk1 in the activation of the Spindle Assembly Checkpoint (SAC) via controlling the protein levels of APC/C activator Cdc20 (Slp1 in S. pombe). The data presented in the manuscript is thorough and convincing. The authors have shown that Pmk1 binds and phosphorylates Slp1, promoting its ubiquitination and subsequent degradation. Since Cdc20 is an activator of APC/C, which promotes anaphase entry, constitutive Pmk1 activation leads to an increased percentage of metaphase-arrested cells. The authors have used genetic and environmental stress conditions to modulate MAP kinase signalling and demonstrate their effect on APC/C activation. This work provides evidence for the role of MAP kinases in cell cycle regulation in S. pombe and opens avenues for exploration of similar regulation in other eukaryotes.

Strengths:

The authors have done a very comprehensive experimental analysis to support their hypothesis. The data is well represented, and including a model in every figure summarizes the data well.

Weaknesses:

As mentioned in the comments, the manuscript does not establish that MAP kinase activity leads to genome stability when cells are subjected to genotoxic stressors. That would establish the importance of this pathway for checkpoint activation.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public Review):

Summary:

This manuscript addresses two main issues:

(i) do MAPKs play an important role in SAC regulation in single-cell organism such as S. pombe?

(ii) what is the nature of their involvement and what are their molecular targets?

The authors have extensively used the cold-sensitive β-tubulin mutant to activate or inactivate SAC employing an arrest-release protocol. Localization of Cdc13 (cyclin B) to the SPBs is used as a readout for the SAC activation or inactivation. The roles of two major MAPK pathways i.e. stress-activated pathway (SAP) and cell integrity pathway (CIP), have been explored in this context (with CIP more extensively than SAP). sty1Δ or pmk1Δ mutants were used to inactivate the SAP or CIP pathways and wis1DD or pek1DD expression was utilized to constitutively activate these pathways, respectively. Lowering of Slp1Cdc20 abundance (by phosphorylation of Slp1-Thr 480) is revealed as the main function of MAPK to augment the robustness of the spindle assembly checkpoint.

Strengths:

The experiments are generally well-conducted, and the results support the interpretations in various sections. The experimental data clearly supports some of the key conclusions:

(1) While inactivation of SAP and CIP compromises SAC-imposed arrest, their constitutive activation delays the release from the SAC-imposed arrest.

(2) CIP signaling, but not SAP signaling, attenuates Slp1Cdc20 levels.

(3) Pmk1 and Cdc20 physically interact and Pmk1-docking sequences in Slp1 (PDSS) are identified and confirmed by mutational/substitution experiments.

(4) Thr480 (and also S76) is identified as the residue phosphorylated by Pmk1. S28 and T31 are identified as Cdk1 phosphorylation sites. These are confirmed by mutational and other related analyses.

(5) Functional aspects of the phosphorylation sites have been elucidated to some extent: (a) Phosphorylation of Slp1-T480 by Pmk1 reduces its abundance thereby augmenting the SAC-induced arrest; (b) S28, T31 (also S59) are phosphorylated by Cdk1; (c) K472 and K479 residues are involved in ubiquitylation of Slp1.

Weaknesses:

(1) Cdc13 localization to SPBs has been used as a readout for SAC activation/inactivation throughout the manuscript. However, the only image showing such localization (Figure 1C) is of poor quality where the Cdc13 localization to SPBs is barely visible. This should be replaced by a better image.

We have replaced those pictures with a new set of representative images, which show clear presence or absence of SPB-localized Cdc13-GFP.

(2) The overlapping error bars in Cdc13-localization data in some figures (for instance Figure 3E and 4H) make the effect of various mutations on SAC activation/inactivation rather marginal. In some of these cases, Western-blotting data support the authors' conclusions better.

We agree that the overlapping error bars may look ambiguous in most figures showing time course curves, this is due to the fact that all these data from a group of strains have to be better presented in a single graph to more directly compare the potential effects. We have been fully aware of the drawback of these figure representations, that is why we always presented the data corresponding two major time points (0 and 50 min after release) from all time course analyses in an alternative way, namely using individual histograms to represent the data from each strain with means of repeats, absolute values, error bars and p values clearly labeled. In particular, the data from time point 0 min can provide important information on the SAC activation efficiency. Generally, we placed those data and graphs in corresponding supplemental figures, such as: Figure 1-figure supplement 1C, Figure 1-figure supplement 2D, Figure 3-figure supplement 3, Figure 4-figure supplement 6B, Figure 5-figure supplement 1, and Figure 6-figure supplement 2.

In addition, as you have noticed, almost all time course data were backed up by our Western blotting data.

(3) This specific point is not really a weakness but rather a loose end:

One of the conclusions of this study is that MAPK (Pmk1) contributes to the robustness of SAC-induced arrest by lowering the abundance of Slp1Cdc20. The authors have used pmk1Δ or constitutively activating the MAPK pathways (Pek1DD) and documented their effect on SAC activation/inactivation dynamics. It is not clear if SAC activation also leads to activation of MAPK pathways for them to contribute to the SAC robustness. To tie this loose end, the author could have checked if the MAPK pathway is also activated under the conditions when SAC is activated. Unless this is shown, one must assume that the authors are attributing the effect they observe to the basal activity of MAPKs.

We agree with your concern. We have followed your suggestion and performed further experiments. Please see our more detailed response to your point #ii(a) in your “Recommendations for the authors”.

(4) This is also a loose end:

The authors show that activation of stress pathways (by addition of KCl for instance) causes phosphorylation-dependent Slp1Cdc20 downregulation (Figure 6) under the SAC-activating condition. Does activation of the stress pathway cause phosphorylation-dependent Slp1Cdc20 downregulation under the non-SAC-activation condition or does it occur only under the SAC-activating condition?

We agree with your concern. We have followed your suggestion and performed further experiments. Please see our more detailed response to your point #ii(b) in your “Recommendations for the authors”.

(5) Although the authors have gone to some length to identify S28 and T31 (also S59) as phosphorylation sites for Cdk1, their functional significance in the context of MAPK involvement is not yet clear. Perhaps it is outside the scope of this study to dig deeper into this aspect more than the authors have.

Based on our data from Mass spectrometry analysis, mutational analysis, in vitro and in vivo kinase assays using phosphorylation site-specific antibodies, we confirmed that at least S28 and T31 are Cdk1 phosphorylation sites. From our time course analysis of these phosphorylation-deficient mutants, it seems the mechanisms of Slp1 activity or protein abundance regulated by Cdk1 or MAPK are quite different. How these two or even more kinases coordinate to control Slp1 activity during APC/C activation is one very interesting issue to be investigated, however, as you have realized, it is indeed beyond the scope of our current study.

(6) In its current state, the Discussion section is quite disjointed. The first section "Involvement of MAPKs in cell cycle regulation" should be in the Introduction section (very briefly, if at all). It certainly does not belong to the Discussion section. In any case, the Discussion section should be more organized with a better flow of arguments/interpretations.

We have re-organized our “Discussion” section. Please see our more detailed response to your point #iii in your “Recommendations for the authors”.

Reviewer #2 (Public Review):

Summary:

This study by Sun et al. presents a role for the S. pombe MAP kinase Pmk1 in the activation of the Spindle Assembly Checkpoint (SAC) via controlling the protein levels of APC/C activator Cdc20 (Slp1 in S. pombe). The data presented in the manuscript is thorough and convincing. The authors have shown that Pmk1 binds and phosphorylates Slp1, promoting its ubiquitination and subsequent degradation. Since Cdc20 is an activator of APC/C, which promotes anaphase entry, constitutive Pmk1 activation leads to an increased percentage of metaphase-arrested cells. The authors have used genetic and environmental stress conditions to modulate MAP kinase signalling and demonstrate their effect on APC/C activation. This work provides evidence for the role of MAP kinases in cell cycle regulation in S. pombe and opens avenues for exploration of similar regulation in other eukaryotes.

Strengths:

The authors have done a very comprehensive experimental analysis to support their hypothesis. The data is well represented, and including a model in every figure summarizes the data well.

Weaknesses:

As mentioned in the comments, the manuscript does not establish that MAP kinase activity leads to genome stability when cells are subjected to genotoxic stressors. That would establish the importance of this pathway for checkpoint activation.

We understand your concern. We have followed your suggestion and performed further experiments to examine whether the absence of Pmk1 causes chromosome segregation defects. Please see our more detailed response to your point #5 in your “Recommendations for the authors”.

Recommendations for the authors:

Reviewing Editor

Please go through the reviews and recommendations and revise the paper accordingly. I think nearly everything is very straightforward and all issues raised by the two expert referees are fully justified. I look forward to seeing an appropriately revised manuscript.

Reviewer #1 (Recommendations For The Authors):

(i) Cdc13 localization to SPBs has been used as a readout for SAC activation/inactivation throughout the manuscript. However, the only image showing such localization (Figure 1C) is of poor quality where the Cdc13 localization to SPBs is barely visible. This should be replaced by a better image.

We have replaced those pictures with a new set of representative images, which show clear presence or absence of SPB-localized Cdc13-GFP.

(ii) I reiterate the loose ends in this manuscript I have mentioned above. If the authors have already conducted these experiments, they should include the results in the manuscript to tighten the story further. (I am not suggesting that the authors must perform these experiments...if they have not).

(a) One of conclusions of this study is that MAPK (Pmk1) contributes to the robustness of SAC-induced arrest by lowering the abundance of Slp1Cdc20. The authors have used pmk1Δ or constitutively activating the MAPK pathways (pek1DD) and documented their effect on SAC activation/inactivation dynamics. It is not clear if SAC activation also leads to activation of MAPK pathways for them to contribute to the SAC robustness. To tie this loose end, the author could have checked if the MAPK pathway is also activated under the conditions when SAC is activated. Unless this is shown, one must assume that the authors are attributing the effect they observe to the basal activity of MAPKs.

Actually, our data shown in Figure 6B demonstrated that SAC activation per se cannot trigger activation of MAPK pathway CIP, because we did not observe any elevated Pmk1 phosphorylation (i.e. Pmk1-P detected by anti-phospho p42/44 antibodies) in nda3-arrested cells (Please see “control” samples in Figure 6B).

To corroborate this observation, we further examined the Pmk1 phosphorylation/activation in Mad2-overexpressing cells, and could not detect elevated Pmk1 phosphorylation. This data again lends support to the notion that SAC activation per se cannot trigger activation of CIP signaling.

We have added our newly obtained result in Figure 6-figure supplement 1 in our revised manuscript.

(b) The authors show that activation of stress pathways (by addition of KCL instance) causes phosphorylation-dependent Slp1Cdc20 downregulation (Figure 6) under the SAC-activating conditions. Does activation of the stress pathway cause phosphorylation-dependent Slp1Cdc20 downregulation under the non-SAC-activation conditions or does it occur only under the SAC-activating condition?

As you suggested, we have constructed cdc25-22 background strains with pmk1+ deleted or expressing Padh11-pek1DD to remove or constitutively activate CIP signaling, respectively. By immunoblotting, we followed the Slp1Cdc20 levels when cells went through mitosis after being released at 25 °C from G2/M-arrest at high temperature. We found that Slp1Cdc20 levels in pek1DD cells were only marginally reduced compared to wild-type cells, whereas we failed to observe any elevated Slp1Cdc20 levels in pmk1Δ cells. These results suggested that CIP signaling only plays a negligible role in influencing Slp1Cdc20 levels under the non-SAC-activation conditions.

We have presented our newly obtained result in Figure 2-figure supplement 1 in our revised manuscript.

(iii) The Discussion section is quite disjointed. The first section "Involvement of MAPKs in cell cycle regulation" should be in the Introduction section (very briefly, if at all). It certainly does not belong to the Discussion section. In any case, the Discussion section should be more organized with a better flow of arguments/interpretations.

Thank you for suggestion on the organization and flow for “Discussion”. We have reorganized our “Discussion” sections and moved the previous “Involvement of MAPKs in cell cycle regulation” to the section “Introduction” and rewrote the corresponding paragraph.

(iv) A minor point in this context:

In the cold-sensitive β-tubulin mutant, growth at 18C causes loss of kinetochore-microtubule attachments as well as the intra-kinetochore tension. Both perturbations individually can lead to the activation of SAC. This study does not distinguish whether MAPK involvement in SAC dynamics is relevant to one perturbation or another or both. It would be pertinent to briefly mention this point in the Discussion section.

As you suggested, we have added two sentences to briefly mention this point in our “Discussion” section.

Reviewer #2 (Recommendations For The Authors):

This study by Sun et al. presents a role for the S. pombe MAP kinase Pmk1 in the activation of the Spindle Assembly Checkpoint (SAC) via controlling the protein levels of APC/C activator Cdc20 (Slp1 in S. pombe). The data presented in the manuscript is thorough and convincing. The authors have shown that Pmk1 binds and phosphorylates Slp1, promoting its ubiquitination and subsequent degradation. Since Cdc20 is an activator of APC/C, which promotes anaphase entry, constitutive Pmk1 activation leads to an increased percentage of metaphase-arrested cells. The authors have used genetic and environmental stress conditions to modulate MAP kinase signalling and demonstrate their effect on APC/C activation. This work provides evidence for the role of MAP kinases in cell cycle regulation in S. pombe and opens avenues for exploration of similar regulation in other eukaryotes.

Although the data largely supports the conclusions, a major addition will be testing whether cells accumulate chromosomal or inheritance defects when MAPK Pmk1 is absent. It will be interesting to know that this mechanism of SAC activation contributes to genome integrity.

Some additions that can improve the manuscript are mentioned below:

(1) In Figure 1, the authors should also test the effect of constitutive activation of Spk1 to rule out the involvement of the PSP pathway.

To meet your curiosity and requirement, we have constructed yeast strains expressing constitutively active byr1DD alleles carrying S214D and T218D point mutations under the control of the adh21 or adh11 promoters (Padh21 or Padh11 in short), i.e. Padh21-6HA-byr1DD and Padh11-6HA-byr1DD, respectively. We examined the expression of these byr1DD alleles by Western blotting, and tested the TBZ sensitivity of these alleles and also checked whether they affect the efficiency of SAC activation or inactivation. Our results showed that constitutive activation of Spk1 by overexpressing Byr1DD does not cause yeast cells to be TBZ-sensitive or affect the efficiency of SAC activation or inactivation.

We have added these new data in Figure 1-figure supplement 2 in our revised manuscript.

(2) The number of analyzed cells (n) should be mentioned in the figure legends in Figure 1D, and all other figure panels should represent similar data in the consequent figures.

We have added the information on sample size for all experiments involving time course analyses.

(3) The authors should also use another arresting mechanism (e.g. nocodazole treatment) and corroborate the result in Figure 1C to rule out any effects due to the mutant.

Figure 1C in our manuscript actually shows our experimental design and not the result. We guess here you asked for alternative strategy to arrest cells at metaphase and confirm our results shown in Figure 1D.

We need to mention that, as a commonly used inhibitor of microtubule polymerization, Nocodazole is very effective in mammalian and human cells and also in budding yeast cells, but not effective at all in wild-type fission yeast cells. It has been found that Nocodazole is only active in fission yeast α- or β-tubulin mutants (please see Umesono, K., et al., J Mol Biol. 168 (2): 271-284 (1983); PMID: 6887245; DOI: 10.1016/s0022-2836(83)80018-7.) or multidrug resistance (MDR) transporter mutants (please see Kawashima, SA, et al., Chemistry & Biology 19, 893–901 (2012); PMID: 22840777; doi: 10.1016/j.chembiol.2012.06.008.). Therefore, this feature of Nocodazole has limited and restricted its routine use as a metaphase arrest or spindle checkpoint activation drug in fission yeast.

Instead, in order to achieve the spindle checkpoint activation and metaphase arrest, we took advantage of a metaphase-arresting mechanism involving Mad2 overexpression, which has been described and used previously (Please see He, X., et al., Proc Natl Acad Sci USA. 94 (15): 7965-70 (1997); PMID: 9223296; DOI: 10.1073/pnas.94.15.7965, and May, K.M., et al., Current Biology, 27(8):1221-1228 (2017); PMID: 28366744; DOI: 10.1016/j.cub.2017.03.013). With this strategy, we could analyze the metaphase-arresting and SAC-activation efficiency by counting cells with short spindles as judged by GFP-Atb2 signals. We compared the frequencies of cells with short spindles in wild-type, pmk1Δ, sty1-T97A, and spk1Δ backgrounds after Mad2 has been induced to overexpress for 18 hours, and found that SAC-activating efficiency was compromised in pmk1Δ and sty1-T97A cells, but not in spk1Δ cells. This data indeed corroborated our result shown in Figure 1D and ruled out possible effects due to the nda3-KM311 mutant.

We have added this new data in Figure 1-figure supplement 3 in our revised manuscript.

(4) It would also be helpful to assess SAC or APC/C activation with another cellular readout in addition to Cdc13-GFP accumulation on SPBs, at least for initial experiments.

Actually, Cdc13-GFP accumulation on SPBs has been routinely used as a reliable cellular readout for SAC or APC/C activation in the field. It was first developed and used by Kevin Hardwick lab in their paper (Vanoosthuyse V and Hardwick KG. Curr Biol. 2009, 19(14):1176-81. PMID: 19592249; doi: 10.1016/j.cub.2009.05.060.). This method was also used in a paper by Meadows JC, et al. (2011) (Dev Cell. 20(6):739-50. PMID: 21664573; doi: 10.1016/j.devcel.2011.05.008.).

In our previous study, we also employed a different strategy to assess SAC inactivation or APC/C activation, in which degradation of nuclear Cut2-GFP was used as a cellular readout (Please see S4 Fig in Bai S, et al., PLoS Genet 18(9): e1010397 (2022); PMID: 36108046; DOI: 10.1371/journal.pgen.1010397.). Cut2 is the securin homologue in S. pombe and therefore also a target of APC/C at anaphase. Our data in the above paper confirmed that the degradation of both nuclear Cut2-GFP and SPB-localized Cdc13-GFP shows similar dynamics when cells are released from metaphase-arrest.

As we described in our response to your comment #3, we employed short spindles visualized by GFP-Atb2 signals as an alternative readout for metaphase-arrest and SAC-activation in cells overexpressing Mad2. We confirmed that SAC-activation efficiency was compromised in pmk1Δ and sty1-T97A cells, but not in spk1Δ cells.

(5) The authors have shown a role for Pmk1 in controlling the activation of APC/C and, hence, cell cycle progression through metaphase to anaphase. One crucial experiment would be to check if pmk1Δ cells show an accumulation of chromosomal aberrations or unequal distribution when subjected to genotoxic stressors. That would implicate a direct importance on Pmk1's role in cell cycle arrest and genome maintenance.

As you suggested, we have constructed cdc25-22 GFP-atb2+ strains with pmk1+ present or deleted, and treated cells with 0.6 M KCl or 2 μg/mL caspofungin to activate MAPKs and checked whether the absence of pmk1 could cause defective chromosome segregation in anaphase cells. Indeed, we found that stressed pmk1Δ cells displayed greatly increased frequency of lagging chromosomes and chromosome mis-segregation at mitotic anaphase compared to similarly treated wild-type cells and also untreated pmk1Δ cells. This new data implicated a direct role for Pmk1 in cell cycle arrest and genome maintenance, especially when cells are exposed to adverse environment.

We have presented this new data as Figure 7 in our revised manuscript.

Typos:

(1) In line 406, "docking" is misspelled as "docing".

Thank you for pointing this out. We have corrected this mistake.

(2) In Figure 6, panel "F" is not marked in the figure.

We mistakenly mentioned and labeled “F” in Figure 6 legend. In our revised manuscript, we have added new results of protein levels of Pmk1 phosphorylation- and ubiquitylation-deficient Slp1Cdc20 mutants upon SAC activation detected by Western blotting in Figure 6-figure supplement 3.

(3) In Figure S1, panel "D" is not marked.

We apologize for our previous wording in our former Figure S1 legend, which was misleading. Actually, there was no panel “D” in Figure S1 (now Figure 1-figure supplement 1). We have rewritten the legend to avoid ambiguity.

Version 2

Abstract

Abstract

Mitotic anaphase onset is a key cellular process tightly regulated by multiple kinases. The involvement of mitogen-activated protein kinases (MAPKs) in this process has been established in Xenopus egg extracts. However, the detailed regulatory cascade remains elusive, and also it is unknown whether the MAPKs-dependent mitotic regulation is evolutionarily conserved in the single cell eukaryotic organism such as fission yeast (Schizosaccharomyces pombe). Here we show that two MAPKs in S. pombe indeed act in concert to restrain anaphase-promoting complex/cyclosome (APC/C) activity upon activation of the spindle assembly checkpoint (SAC). One MAPK, Pmk1, binds and phosphorylates Slp1Cdc20, the co-activator of APC/C. Phosphorylation of Slp1Cdc20 by Pmk1, but not by Cdk1, promotes its subsequent ubiquitylation and degradation. Intriguingly, Pmk1-mediated phosphorylation event is also required to sustain SAC under environmental stress. Thus, our study establishes a new underlying molecular mechanism of negative regulation of APC/C by MAPK upon stress stimuli, and provides an unappreciated framework for regulation of anaphase entry in eukaryotic cells.

Introduction

The evolutionarily conserved mitogen-activated protein kinase (MAPK) signaling pathways regulate multiple cellular functions in eukaryotic organisms in response to a wide variety of environmental cues (Plotnikov et al., 2011). However, different MAPK pathways have been evolved in an organism to integrate diverse signals and fulfill different regulations on various effectors (Cansado et al., 2021; Ronkina and Gaestel, 2022). This makes specifying the MAPKs and substrates involved in a specific physiological process rather complex and challenging.

The fission yeast Schizosaccharomyces pombe has three MAPK-signaling cascades: the pheromone signaling pathway (PSP), the stress-activated pathway (SAP) and the cell integrity pathway (CIP), with Spk1, Sty1 or Pmk1 as MAPK respectively (Figure 1A) (Cansado et al., 2021; Perez and Cansado, 2010). So far, the only known cell-cycle control stages linked to MAPKs in fission yeast are at G2/M transition and during cytokinesis (Gomez-Gil et al., 2020; Petersen and Hagan, 2005; Petersen and Nurse, 2007). The p38 MAPK family member Sty1 either arrests or promotes mitotic commitment in unperturbed, stressed or resumed cell cycles after stress depending on the downstream kinases it associates with. Sty1 can phosphorylate kinase Srk1 or Polo kinase Plo1 to negatively or positively regulate Cdc25 phosphatase and thus the onset of mitosis, respectively (Lopez-Aviles et al., 2005; Lopez-Aviles et al., 2008; Petersen and Hagan, 2005; Petersen and Nurse, 2007; Smith et al., 2002). Recent studies have also demonstrated that Sty1 and Pmk1 can pose negative effect on assembly of contractile actomyosin ring (CAR) and cytokinesis in response to actin cytoskeleton damage or cell wall stress (e.g. treatment with blankophor, caspofungin and caffeine) (Edreira et al., 2020; Gomez-Gil et al., 2020), though the detailed mechanisms have not been elucidated.

pmk1Δ and sty1-T97A mutants display spindle checkpoint activation defects, while pek1DD and wis1DD mutants are defective in checkpoint silencing.

(A) Schematic of core modules of three Schizosaccharomyces pombe mitogen-activated protein kinase (MAPK) signaling pathways. Each cascade consists of three core kinases (MAP kinase kinase kinase (MAPKKK), MAP kinase kinase (MAPKK), and MAPK). CIP, the cell integrity pathway; SAP, the stress-activated pathway; PSP, pheromone signaling pathway. These pathways can be disrupted by pmk1Δ, sty1Δ or ATP analogue-sensitive mutation sty1-T97A or spk1Δ, and constitutively activated by mutations in MAPKKs Pek1 (pek1DD, pek1-S234D;T238D), Wis1 (wis1DD, wis1-S469D;T473D) or Byr1 (byr1DD, byr1-S214D;T218D), respectively.

(B) Serial dilution assay on TBZ sensitivity of all MAPK deletion mutants and pek1DD- or wis1DD-overexpressing mutants. mad2Δ is a positive control. Note Padh11 is a stronger version of Padh21promoter.

(C) Schematic depiction of the experiment design for time-course analyses on SAC or APC/C activation. nda3-KM311 cells carrying Cdc13-GFP were grown, synchronized with HU and treated at 18 °C for 6 hours to activate SAC, and finally shifted back to the permissive temperature 30 °C. Samples were collected at 10 min intervals and subjected to microscopy analyses. Example pictures of cells with Cdc13-GFP signals enriched or disappeared at spindle pole bodies (SPBs) are shown. Scale bar, 5 μm.

(D) Time-course analyses of SAC activation and inactivation in nda3-KM311 cdc13-GFP strains with indicated genotypes. sty1-T97A was inactivated by 5μM 3-BrB-PP1. For each time point, ≥300 cells were counted for every sample. The experiment was repeated ≥3 times and error bars indicate mean ± standard deviation.

In late mitosis, the timely polyubiquitylation and subsequent degradation of securin and cyclin B by anaphase-promoting complex/cyclosome (APC/C) play a critical role for anaphase onset and chromosome segregation (Peters, 2006; Sullivan and Morgan, 2007; Yamano, 2019). Given the essential role of APC/C in triggering chromosome segregation, it is not surprising that APC/C is a key molecular target of the spindle assembly checkpoint (SAC), which is an intricate surveillance mechanism that prolongs mitosis until all chromosomes achieve correct bipolar attachments to spindle microtubules (McAinsh and Kops, 2023; Murray, 2011; Musacchio, 2015). Previous studies have revealed that the active MAPK is required for spindle checkpoint activation and APC/C inhibition in Xenopus egg extracts or tadpole cells (Chen, 2004; Chung and Chen, 2003; Minshull et al., 1994; Takenaka et al., 1997; Zhao and Chen, 2006). These studies suggested that phosphorylation of several components of SAC, including Cdc20, Mps1 and Bub1, by MAPK serves as regulatory signals to block APC/C activation upon SAC activation (Chen, 2004; Chung and Chen, 2003; Zhao and Chen, 2006). However, all these studies used either the anti-MAPK antibodies or small molecule inhibitor UO126 to inhibit MAPK or MAPK-specific phosphatase MKP-1 to antagonize MAPK activity, these reagents could not distinguish the contributions of different MAPKs with high specificity. Thus, those seemingly defined mechanisms might suffer the ambiguity due to potential specificity issues.

In S. pombe, whether MAPKs play key roles in regulation of the APC/C and spindle checkpoint remains largely unexplored. Intriguingly, all genes encoding the major components of three MAPK-signaling cascades in fission yeast are non-essential (Kim et al., 2010), this provides the possibility for investigations on their potential functions in a genetic background with complete deletion of relevant genes. In this work, we set out to directly examine the requirement of the fission yeast MAPKs in mitotic APC/C activation and anaphase entry. Our results uncovered a previously unappreciated mechanism in which Pmk1, the MAPK of CIP, phosphorylates the APC/C activator Slp1 (the Cdc20 homologue in S. pombe) upon SAC activation. Phosphorylation of Slp1Cdc20 results in its ubiquitylation and degradation and thus subsequently impedes APC/C activation and anaphase entry. This mechanism also operates in response to spindle defects under environmental stress. Therefore, our finding extends cell-cycle control stages linked to MAPKs from previously recognized G2/M transition and cytokinesis to anaphase entry and mitotic exit in fission yeast. It will also be interesting to address whether this MAPK-mediated Cdc20 regulation exists as a general mechanism for balancing spindle checkpoint activity in higher eukaryotes.

Results

SAP and CIP signaling components are required for spindle checkpoint activity

As the first step to examine the possible requirement of the fission yeast MAPKs in mitotic progression, we tested the sensitivity of deletion mutants of three fission yeast MAPKs to microtubule-destabilizing drug thiabendazole (TBZ), which has been routinely used as an indicator of defective kinetochore or spindle checkpoint (Akera et al., 2015; Saitoh et al., 1997). We noticed that pmk1Δ and sty1Δ but not spk1Δ cells were extremely sensitive to TBZ, and constitutively activated CIP or SAP signaling by overexpressing Pek1DD (Pek1-S234D;T238D) (Sugiura et al., 1999) or Wis1DD (Wis1-S469D;T473D) (Shiozaki et al., 1998) respectively, but not activated PSP signaling by overexpressing Byr1DD (Byr1-S214D;T218D) (Ozoe et al., 2002), also caused yeast cells to be sensitive to TBZ (Figure 1B; Figure 1-figure supplement 1A and Figure 1-figure supplement 2A, B). These observations indicated that Pmk1 and Sty1 pathways might be involved in mitosis-related processes.

To extend our analysis and precisely quantify whether the SAC is properly activated and maintained in MAPK mutants, we adopted one well-established assay using the cold-sensitive β-tubulin mutant nda3-KM311, which compromises the kinetochore-spindle microtubule attachment upon cold treatment (Hiraoka et al., 1984), to analyze the ability of cells to arrest in mitosis in the absence of spindle microtubules and enter anaphase when spindle microtubules are reassembled (Bai et al., 2022; May et al., 2017; Vanoosthuyse and Hardwick, 2009). In this assay, the SAC was first robustly activated by the nda3-KM311 mutant at 18 °C and then inactivated simply by shift mitotically arrested cells back to permissive temperature (30 °C) (Figures 1C). Because Cdc13 (cyclin B in S. pombe) localizes to the spindle pole bodies (SPBs) in early mitosis and should be degraded by APC/C to promote metaphase-anaphase transition, accumulation of Cdc13-GFP on SPBs after cold treatment serves as the read-out of the SAC activation, and the disappearance rate of Cdc13-GFP spot under permissive temperature reflects the SAC inactivation efficacy (Figures 1C). It is noteworthy that the ATP analogue-sensitive mutant sty1-T97A (i.e. sty1-as2) (Zuin et al., 2010), instead of sty1Δ deletion mutant, was used in this assay. That was due to the G2 delay that has been demonstrated in the sty1Δ deletion mutant (Shiozaki and Russell, 1995a), which could interfere mitotic timing for further analysis in this assay. We found that inactivation of CIP or SAP signaling by pmk1Δ or sty1-T97A compromised full SAC activation, as these two mutants had only less than 60% of cells with Cdc13-GFP on SPBs after 6 hours of cold treatment, whereas the percentage in wild-type and spk1Δ cells after the same treatment was roughly 90% (Figure 1D; Figure 1-figure supplement 1C). Consistently, the disappearance of Cdc13-GFP from SPBs in pek1DDand wis1DD but not byr1DD mutant cells released from metaphase-arrest was severely delayed (Figure 1D; Figure 1-figure supplement 1B and Figure 1-figure supplement 2C, D), suggesting delayed SAC inactivation and anaphase onset after release from spindle checkpoint arrest when CIP or SAP signaling is constitutively activated.

To corroborate our above observations of negative effect of CIP or SAP signaling on SAC strength, we took advantage of a different spindle checkpoint activation and metaphase-arresting mechanism involving Mad2 overexpression (May et al., 2017). After Pnmt1-mad2+ was induced to overexpress for 18 hours, we counted cells with short spindles visualized by GFP-Atb2 as indicative of metaphase-arrest and SAC-activation (Figure 1-figure supplement 3A). We found that the ability of Mad2-overexpressing cells to arrest in mitosis was clearly compromised in pmk1Δ and sty1-T97A mutants, but not in spk1Δ mutant (Figure 1-figure supplement 3B).

All above data suggested that MAPK pathways are evolutionarily conserved in the fission yeast to participate in spindle checkpoint activation and maintenance, of which the CIP and SAP signaling pathways, but not the Spk1 pathway, are actively involved.

CIP and SAP signaling attenuate protein levels of Slp1Cdc20 and facilitate association of MCC with APC/C respectively

The mitotic checkpoint complex (MCC, consisting of Mad3-Mad2-Slp1Cdc20 in S. pombe) is the most potent inhibitor of the APC/C prior to anaphase (Kapanidou et al., 2017; Primorac and Musacchio, 2013; Sudakin et al., 2001). Previous study has revealed that Xenopus MAPK phosphorylates Cdc20 and facilitates its binding by Mad2, BubR1 (Mad3 in S. pombe) and Bub3 to form a tight MCC to prevent Cdc20 from activating the APC/C (Chung and Chen, 2003). In fission yeast, the key SAC components Mad2 and Mad3 and one molecule of Slp1Cdc20 form MCC which binds to APC/C through another molecule of Slp1Cdc20 upon checkpoint arrest, and the recovery from mitotic arrest accompanies the loss of MCC-APC/C binding (May et al., 2017; Sczaniecka et al., 2008; Sewart and Hauf, 2017; Vanoosthuyse and Hardwick, 2009).

Since we found that constitutive activation of CIP or SAP signaling in fission yeast by pek1DD or wis1DD mutations caused prolonged SAC activation, we were suspicious that these two MAPK pathways might also execute their functions cooperatively through a similar mechanism to that in Xenopus. To test this possibility, we analyzed the MCC occupancy on APC/C by immunoprecipitations of the APC/C subunit Lid1 (Yoon et al., 2002) in pek1DDand wis1DD cells arrested by activated checkpoint. Compared to wild-type, more MCC components (Mad2 plus Mad3 and Slp1Cdc20) were co-immunoprecipitated in wis1DD cells, while less of them were co-immunoprecipitated in pek1DD cells (Figure 2A), suggesting that activation of the SAP but not the CIP pathway enhanced the association of MCC with APC/C.

Upon spindle checkpoint activation, Slp1Cdc20 levels are reduced in pek1DD and the MCC-APC/C association is enhanced in wis1DD cells compared to wild-type cells.

(A) Co-immunoprecipitation analysis on MCC-APC/C association. Cells with indicated genotypes were grown at 30 °C to mid-log phase and arrested at 18 °C for 6 hours. Lid1-TAP was immunoprecipitated and associated Mad2, Mad3 and Slp1Cdc20 were detected by immunoblotting. The amount of co-immunoprecipitated Mad2, Mad3 and Slp1Cdc20 was quantified by being normalized to those of total immunoprecipitated Lid1 in each sample, with the relative ratio between Mad2-GFP plus Mad3-GFP or Slp1Cdc20 and Lid1-TAP in wild-type sample set as 1.0. Blots are representative of three independent experiments. p values were calculated against wild-type cells.

(B, C) Immunoblot analysis of Slp1Cdc20 abundance in nda3-KM311 cells treated at 18 °C for 6 hr. Slp1Cdc20 levels were quantified with the relative ratio between Slp1Cdc20 and Cdc2 in wild-type strain set as 1.0. Phosphorylated Pmk1 (Pmk1-P) or phosphorylated Sty1 (Sty1-P) in (A) were detected using anti-phospho p42/44 and anti-phospho p38 antibodies and represents activated CIP or SAP signaling, respectively. sty1-T97A was inactivated by 5μM 3-BrB-PP1. Blots shown are the representative of three independent experiments. p values were calculated against wild-type cells.

(D) RT-qPCR analysis of mRNA levels of slp1+. Cells with indicated genotypes were grown and treated as in (A-C) before RNA extraction. The relative fold-change (slp1+/act1+) in mRNA expression was calculated with that in wild-type cells being normalized to 1.0. Note mRNA level of slp1+ in pek1DD mutant is not decreased. Error bars indicate mean ± standard deviation of three independent experiments. Two-tailed unpaired t-test was used to derive p values.

(E) Schematic summary of the negative effect of activated CIP and SAP signaling on APC/C activation based on primary phenotype characterization of pmk1Δ, sty1-T97A, pek1DD and wis1DD mutants.

In the course of the above immunoprecipitations, we noticed the decrease of co-immunoprecipitated Mad2 and Mad3 in pek1DD cells, which was largely out of expectation. Further analysis of the inputs of these immunoprecipitations showed that Slp1Cdc20 levels were significantly reduced in pek1DD but not wis1DD cells (Figure 2A). Our immunoblotting analyses using the total lysate of pek1DD or wis1DD cells arrested at metaphase also confirmed these results (Figure 2B). Consistently, the Slp1Cdc20 level was significantly increased in pmk1Δ and pmk1Δ sty1-T79A cells but not in sty1-T79A cells (Figure 2C). The elevated Slp1Cdc20 level in pmk1Δ cells is similar to that we observed in cells harboring two copies of slp1+, which leads to artificially increased Slp1Cdc20 level and also causes a SAC strength defect (Bai et al., 2022). It is noteworthy that attenuated levels of Slp1Cdc20 in pek1DD cells was not due to decreased transcription of slp1+ gene (Figure 2D). These results collectively revealed a negative effect of CIP but not SAP signaling on Slp1Cdc20 levels upon SAC activation, which potentially results in the weakened MCC-APC/C interaction detected in CIP mutants (Figure 2A).

Next, we also tested whether activation of the CIP signaling causes downregulation of Slp1Cdc20 in normally growing mitotic cells, which were enriched by release of cdc25-22 background cells from G2/M-arrest to 25 °C for 60-90 minutes (Figure 2-figure supplement 1A, B). Intriguingly, we could detect only marginally reduced Slp1Cdc20 levels in pek1DD cells, which was consistent with the slightly retarded anaphase entry in those cells, but we failed to observe any elevated Slp1Cdc20 levels in pmk1Δ cells (Figure 2-figure supplement 1C). These results suggested that CIP signaling only plays a negligible role in influencing Slp1Cdc20 levels when SAC is not activated.

Together, our above data suggested a new mechanism involves division of labor between the two fission yeast MAPK pathways CIP (Pek1-Pmk1) and SAP (Wis1-Sty1) upon SAC activation, which is required in concert to lower Slp1Cdc20 levels and enhance MCC affinity for APC/C to strongly inhibit APC/C activity (Figure 2E). In this study, we only focused on characterizing the role of the CIP/Pmk1 in regulating SAC inactivation and anaphase onset hereafter.

Pmk1 directly binds Slp1Cdc20 to attenuate its levels upon SAC activation

To gain further insight into how Slp1Cdc20 levels are modulated by Pmk1, we examined the potential interaction of Pmk1 with Slp1Cdc20. Through incubating MBP-tagged Slp1Cdc20 with metaphase-arrested yeast cell lysates, we found that Pmk1 expressed from yeast cells could be pull-downed by Slp1Cdc20 (Figure 3A).

Pmk1 directly binds Slp1Cdc20 to attenuate its levels upon spindle checkpoint activation.

(A) In vitro binding assay using bacterially expressed MBP-Slp1Cdc20 and yeast lysates prepared from nda3-KM311 pmk1-HA-6His cells arrested at 18 °C for 6 hr. Note that weak band detected in MBP sample was due to unspecific background binding to amylose beads. Coomassie blue staining shows inputs for MBP and MBP-Slp1Cdc20.

(B) Schematic depiction of the S. pombe Slp1Cdc20 protein structure with the positions of two confirmed basic-residue patches mediating Slp1Cdc20-Pmk1 association indicated by pink bars. Alignment highlights the conservation of basic-residue patches within 4 Schizosaccharomyces species. The deduced Pmk1-docking motifs and different versions of motif mutations are shown. MIM, Mad2-interaction motif; IR, isoleucine-arginine tail.

(C) In vitro GST pull-down assays with bacterially expressed recombinant GST-Pmk1 and MBP-fusions of wild-type Slp1Cdc20 or Slp1Cdc20 mutants harboring Pmk1-docking motif mutations. An aliquot of the same amount of MBP-Slp1Cdc20 as that added in each GST pull-down reactions was immobilized by amylose resin as the input control. Asterisks indicate unspecific or degraded protein bands.

(D) Immunoblot analysis of Slp1Cdc20 abundance in nda3-KM311 cells with indicated genotypes treated at 18 °C for 6 hr. Slp1Cdc20 levels were quantified as in Figure 2B and 2C. The experiment was repeated 3 times. The mean value for each sample was calculated, and p values were calculated against wild-type or pek1DD cells.

(E) Time-course analyses of SAC activation and inactivation in nda3-KM311 cdc13-GFP strains with indicated genotypes. For each time point, ≥300 cells were counted for every sample. The experiment was repeated 3 times and the mean value for each sample was calculated as in Figure 1D.

(F) Schematic summarizing the negative effect of Pmk1-Slp1Cdc20 association on Slp1Cdc20 abundance and APC/C activation.

In previous reports, MAPK-docking sites, which are important for interaction with MAPKs, have been identified at the N termini of different MAPKKs from yeast to humans with a notable feature of a cluster of at least two basic residues (mainly Lys/K and Arg/R) separated by a spacer of 2-6 residues from a hydrophobic-X-hydrophobic sequence (Bardwell et al., 2001; Bardwell and Thorner, 1996). Thus, we attempted to identify the potential key residues in Slp1Cdc20 that mediate the Pmk1-Slp1Cdc20 interaction. By visual scanning of Slp1Cdc20 sequence, we noticed five basic-residue patches which are also present in several fungal species and loosely resemble the consensus sequence of the known MAPK-docking sites, and four of them are within N-terminal portion of the protein (Figure 3B and Figure 3-figure supplement 1). However, only mutating either of the two most N-terminal basic-residue patches (19-KKR-21 and 47-KR-48) to glutamates (Glu, E) or alanines (Ala, A) in Slp1Cdc20 resulted in decreased amount of Slp1Cdc20 pulled down by Pmk1-GST in vitro using the bacterially expressed recombinant proteins, and mutating both patches or mutating only two arginine residues (R21 and R48) in those patches was sufficient to completely disrupt their interaction (Figure 3C and Figure 3-figure supplement 2). Importantly, mutations of these Pmk1-docking sequences in Slp1 (PDSS for short) to alanines mimicked pmk1Δ mutant in both elevated Slp1Cdc20 protein levels (Figure 3D) and defective activation and maintenance of the SAC upon nda3-mediated checkpoint arrest (Figure 3E and Figure 3-figure supplement 3). Also, similar to removal of Pek1 downstream MAPK Pmk1, the lowered Slp1Cdc20 protein levels and delayed SAC inactivation in pek1DD mutant could be relieved by slp1(DS-5A) mutant, which is deficient in Pmk1-docking (Figure 3D, 3E and Figure 3-figure supplement 3).

Previously, one conserved non-canonical Pmk1-docking sequence which bears positively charged residues and IYT motif has been identified in MAPK phosphatase Pmp1 (Sacristan-Reviriego et al., 2014). Very strikingly, insertion of a 15 amino acid sequence containing the Pmk1-docking sequence in Pmp1 (or PDSP for short) right in front of the first basic-residue patch in Slp1(DS-5A) mutant protein restored its interaction with Pmk1 (Figure 3B and 3C). And this artificially forced Pmk1-Slp1Cdc20interaction via PDSP was able to lower Slp1Cdc20 protein levels and restore higher percentage of metaphase arrest upon SAC activation in slp1(DS-5A) mutant, which is Pmk1-docking-deficient, although the rescuing effect was less efficient than the presence of genuine PDSS (Figure 3D, 3E and Figure 3-figure supplement 3). These data firmly supported the idea that Pmk1 binds Slp1Cdc20 to downregulate its levels upon SAC activation.

Collectively, these results demonstrated that Pmk1 directly binds to Slp1Cdc20 through two short N-terminal basic patches serving as the major docking sites for Pmk1, and the physical interaction of Pmk1-Slp1Cdc20 is involved in attenuation of Slp1Cdc20 abundance upon SAC activation (Figure 3F).

Pmk1 synergizes with Cdk1 to phosphorylate Slp1Cdc20

The association of Pmk1 with Slp1Cdc20 and the negative effect of constitutively activated CIP signaling on SAC inactivation relies on Pmk1 kinase suggested that Slp1Cdc20 may be a direct substrate of Pmk1. To test this possibility, we performed in vitro phosphorylation reactions and found that bacterially expressed recombinant GST-Slp1Cdc20 could be efficiently phosphorylated in the presence of Pmk1-HA-His purified from yeast cells (Figure 4A).

Pmk1 synergizes with Cdk1 to phosphorylate and reduce Slp1Cdc20 abundance.

(A) Non-radioactive in vitro phosphorylation assays with bacterially expressed recombinant GST-Slp1Cdc20 and Pmk1-HA-His purified from yeast cells. The incorporation of the thiophosphate group was determined using anti-thiophosphate ester antibodies (anti-thioP) as indicative of phosphorylation. Note that the presence or absence of GST-Pek1DD does not affect Slp1Cdc20 phosphorylation efficiency. A known Pmk1 substrate Atf1 was used as a positive control (Atf1-P). Asterisks indicate bands corresponding to unspecific or likely degraded proteins.

(B) Summary of mass spectrometry data on Slp1Cdc20 phosphorylation and ubiquitylation in vivo in nda3-KM311-arrested cells. Red arrows and filled blue circles denote all detected phosphorylated or ubiquitylated sites respectively, while gray arrows and unfilled circles indicate the absence of phosphorylation or ubiquitylation in some of these sites respectively. Alignment highlights the conservation of most of the detected phosphorylation and ubiquitylation sites within 4 Schizosaccharomyces species.

(C) In vitro phosphorylation assays with bacterially expressed recombinant GST-fusion of Slp1Cdc20 fragment (456-488aa), GST-Pmk1 and GST-Pek1DD. The reactions were blotted with pT480 antibodies. Asterisks indicate bands corresponding to unspecific or likely degraded proteins.

(D) Immunoblot detection of Slp1Cdc20 phosphorylation at T480 in vivo. GST-slp1(456-488aa) was purified from mts3-1 cells with indicated genotypes arrested at 36 [for 3.5 hours, and detected with anti-GST and anti-pThr480 antibodies. Note that pThr480 is absent in pmk1Δ cells, and enhanced in pek1DD cells relative to that in wild type cells.

(E) In vitro phosphorylation assays with bacterially expressed recombinant MBP-fusion of Slp1Cdc20 fragment (1-190aa) and Cdc13 (cyclin B)-containing Cdk1 complexes purified from metaphase-arrested nda3-KM311 yeast cells. 1-NM-PP1 was added as inhibitor for analogue-sensitive Cdc2-as. The reactions were blotted with pS28/pT31 antibodies. Asterisks indicate bands corresponding to unspecific or likely degraded proteins.

(F) Immunoblot detection of Slp1Cdc20 phosphorylation at S28/T31 in vivo. Apc15-13myc was immunoprecipitated from nda3-KM311 cells treated at 18 °C for 6 hr and the samples were blotted with pS28/pT31 antibodies. One IP sample from wild type background was treated with λ-phosphatase. For cdc2-asM17 cells, 1-NM-PP1 was added to inactivate Cdc2 during culturing.

(G) Immunoblot analysis of Slp1Cdc20 abundance in nda3-KM311 cells treated at 18 °C for 6 hr. Slp1Cdc20 levels were quantified as in Figure 2B and 2C. The experiment was repeated 3 times. The mean value for each sample was calculated, and p values were calculated against wild-type or pek1DD cells.

(H) Time-course analyses of SAC activation and inactivation in nda3-KM311 cdc13-GFP strains with indicated genotypes. For each time point, ≥300 cells were counted for every sample. The experiment was repeated 3 times and the mean value and p value for each sample were calculated as in Figure 1D.

(I) Schematic summarizing the negative effect of Slp1Cdc20 phosphorylation by Pmk1 on its abundance and APC/C activation.

To identify potential phosphorylation sites in vivo, we purified Apc15-GFP from wild type, pmk1Δ and pek1DD cells arrested by activated checkpoint with intention that the Slp1Cdc20-containing APC/C complexes could be captured (Figure 4-figure supplement 1A and 1B). Purifications were performed in the presence of phosphatase inhibitors to enrich for phosphopeptides. Mass spectrometry analysis revealed that the phosphorylation of Thr480 residue at the most C-terminus of Slp1Cdc20 was dependent on the presence of Pmk1, and this modification was present in pek1DD but not in wild type cells (Figure 4B, Figure 4-figure supplement 1C-E and Figure 4-figure supplement 2), indicating that phosphorylation of Thr480 is very likely triggered by CIP signaling. We raised phospho-specific antibodies against pT480, it could recognize Slp1Cdc20 fragment phosphorylated by Pmk1 in vitro and mutation of Thr480 to alanine (T480A) abolished site-specific phosphorylation (Figure 4C). We also confirmed that endogenous Slp1Cdc20 was indeed phosphorylated at Thr480 in metaphase-arrested mts3-1 cells, which was enhanced in pek1DD mutant and diminished in pmk1Δ mutant (Figure 4D). Furthermore, we examined how in vivo Slp1Cdc20 phosphorylation at Thr480 is affected by forced targeting of Slp1Cdc20 to Pmk1. By taking advantage of the high binding affinity of GFP-binding protein (GBP) to GFP and its variants (Chen et al., 2017; Rothbauer et al., 2006), we constructed strains ectopically expressing Slp1(1-60aa)-mEGFP-2xNLS-GST-Slp1(456-488aa) and Pmk1-GBP-mCherry, which could bring about artificial tethering of Slp1Cdc20 fragments to Pmk1 inside nuclei (Figure 4-figure supplement 3A, B). Immunoblot detection of Slp1Cdc20 phosphorylation at T480 in vivo demonstrated that indeed the presence of Pmk1-GBP-mCherry could increase pThr480 levels of mEGFP-fusion of Slp1Cdc20 fragments, and simultaneously overexpressing Pek1DD from Padh11-6xHA-pek1DD further enhanced the effect (Figure 4-figure supplement 3C). Thus, our results established Thr480 as a major Pmk1 phosphorylation site in vivo upon SAC activation.

In addition to pT480, our mass spectrometry analysis also identified Ser28, Thr31 and Ser59 as Pmk1-independent and Ser76 as Pmk1-dependent phosphorylation sites (Figure 4B, Figure 4-figure supplement 1C-E and Figure 4-figure supplement 2). All these four residues are located within the N-terminal unstructured region, fit with the Cdk1 core motif (SP or TP), and more importantly, they are conserved only in four Schizosaccharomyces species and several other fungal species but not in higher eukaryotes (Figure 4B, Figure 3-figure supplement 1B and Figure 4-figure supplement 4). Interestingly, they were also proposed as Cdk1 phosphorylation sites based on quantitative phosphoproteomic analysis in a recent study (Swaffer et al., 2018) (Figure 4B). By using the phospho-specific antibodies against both pS28 and pT31, we could detect the phosphorylation of Ser28 and Thr31 in vitro by Cdc13 (cyclin B)-containing Cdk1 complexes purified from metaphase-arrested yeast cells, which was diminished when the kinase activity of ATP analogue-sensitive Cdk1 [Cdc2-as1(F84G)] was inhibited by 1-NM-PP1 or when Slp1(S28A;T31A) was added as the substrate in in vitro kinase reactions (Figure 4E). Next, we examined whether Ser28 and Thr31 were phosphorylated in vivo using immunoblotting with phospho-specific antibodies to directly detect Slp1Cdc20, which was co-immunoprecipitated with Apc15 from nda3-arrested yeast cells. We observed phosphorylation of Ser28/Thr31 in wild type cells, but it disappeared following phosphatase treatment of wild type samples after immunoprecipitation, 1-NM-PP1 treatment of cdc2-as1 cells before being collected for immunoprecipitation, or when cells expressing Slp1(S28A;T31A) were used (Figure 4F). Thus, Slp1Cdc20 is also phosphorylated by Cdk1 at least at Ser28 and Thr31 in fission yeast mitotic cells.

We wondered whether Slp1Cdc20 phosphorylation by Pmk1 affects its stability and APC/C activation. Indeed, phospho-deficient Slp1T480A mutant showed elevated Slp1Cdc20 levels in both metaphase-arrested and asynchronously grown cells (Figure 4G and Figure 4-figure supplement 5) and compromised SAC activation compared to wild type cells, likely due to more efficient APC/C activation (Figure 4H and Figure 4-figure supplement 6). Furthermore, Slp1T480A mutant also reversed the inhibitory effect of pek1DD on Slp1Cdc20 stability and APC/C activation after release from SAC-mediated arrest (Figure 4G and 4H).

Ser76 was identified as a Pmk1-dependent phosphorylation residue in our mass spectrometry analysis, but it was considered as a Cdk1 targeting site in an earlier study (Swaffer et al., 2018). It is fairly possible that phosphorylation of Ser76 is promoted by both Cdk1 and Pmk1, which is conceivable as CDKs and MAPKs recognize the same core motif (SP/TP) (Pinna and Ruzzene, 1996), and crosstalk between CDK and MAPK has recently been reported for pheromone signaling in budding yeast (Repetto et al., 2018). We tested how the double phosphorylation sites (S76A;T480A) mutant affects Slp1Cdc20 levels or APC/C activity, and observed that the protein levels of Slp1S76A;T480A were further elevated and SAC activation efficiency in this mutant was further lowered compared to Slp1T480A mutant (Figure 4G and 4H). Of note, Slp1T480A and Slp1S76A;T480A mutant proteins became easily detected in interphase cells (Figure 4-figure supplement 5), this is very different from wild-type Slp1 which normally accumulates exclusively in mitosis (Yamada et al., 2000).

We further tested whether replacement of other identified phospho-serine/threonine residues with alanines affects Slp1Cdc20 protein levels and APC/C activity. We constructed strains carrying combinations of mutations of S28A, T31A and S59A with S76A, and also with confirmed Pmk1 phosphorylation-deficient T480A, and found that mutations of S28A, T31A and S59A did not affect Slp1Cdc20 protein levels, though they could also profoundly compromise SAC activation efficiency similarly to S76A;T480A mutant (Figure 4G, 4H, Figure 4-figure supplement 5, and Figure 4-figure supplement 6). We noticed that the presence of mutations of S28A, T31A and S59A also abolished the elevated Slp1Cdc20 protein levels conferred by S76A;T480A mutations (Figure 4G and Figure 4-figure supplement 5). Thus, phosphorylation of S28, T31 and S59 targeted by Cdk1 most likely influences APC/C activation through distinct mechanism from Pmk1-mediated S76/T480 phosphorylation.

Taken together, our data suggested that activated Pmk1 phosphorylates Slp1Cdc20 in vivo, which leads to lowered Slp1Cdc20 levels, delayed SAC inactivation and APC/C activation as we observed in pek1DDcells. Also, Pmk1 can join Cdk1 to enhance the multi-site phosphorylation of their shared substrate Slp1Cdc20 and collaboratively dampen APC/C activity.

Pmk1- but not Cdk1-mediated Slp1Cdc20 phosphorylation promotes its ubiquitylation

Our above data support the notion that the direct binding of Slp1Cdc20 by Pmk1 mediates its phosphorylation, which possibly results its subsequent rapid turnover. It has been shown in both human and fission yeast cells that Apc15, one of the conserved components of APC/C, promotes Cdc20/Slp1Cdc20 autoubiquitylation and its turnover by APC/C (Mansfeld et al., 2011; May et al., 2017; Sewart and Hauf, 2017; Uzunova et al., 2012). The increased Slp1Cdc20 levels in pmk1Δ cells (Figure 2C) is reminiscent of the previous observations in apc15Δ mutant, which also shows elevated levels of Slp1Cdc20 (May et al., 2017; Sewart and Hauf, 2017). We reasoned that the absence of Apc15 may reverse the negative effect of pek1DD on Slp1Cdc20 levels and spindle checkpoint inactivation and APC/C activation. As expected, we indeed observed that deletion of apc15 recovered Slp1Cdc20 levels in pek1DD cells (Figure 5A), and also lowered the percentage of cells with Cdc13-GFP retained on SPBs in nda3-KM311 pek1DD cells after cold treatment (Figure 5B and Figure 5-figure supplement 1). These observations are consistent with previous reports showing the relative abundance between checkpoint proteins and the checkpoint target Slp1Cdc20 is an important determinant of checkpoint robustness (Heinrich et al., 2013) and apc15Δ mutant is spindle checkpoint-defective (May et al., 2017).

Pmk1- but not Cdk1-mediated Slp1Cdc20 phosphorylation promotes its ubiquitylation.

(A) Immunoblot analysis of Slp1Cdc20 abundance in apc15+ or apc15Δ background cells with indicated genotypes after being treated at 18 °C for 6 hr. Slp1Cdc20 levels were quantified as in Figure 2B. The experiment was repeated 3 times. The mean value for each sample was calculated, and p values were calculated against wild-type or pek1DD cells.

(B) Time-course analyses of SAC activation and inactivation in nda3-KM311 cdc13-GFP strains with indicated genotypes. The experiments were performed and analyzed as in Figure 1D.

(C) Immunoblot analysis of Slp1Cdc20 abundance in K472R, K479R or K472R/K479R mutants. The experiment was repeated 3 times. The mean value for each sample was calculated, and p values were calculated against wild-type or pek1DDcells.

(D) Time-course analyses of SAC activation and inactivation in nda3-KM311 cdc13-GFP strains with K472R, K479R or K472R/K479R mutations. The experiments were performed as in (B).

(E) Schematic depiction of affinity pull-down assays using TUBE (Tandem Ubiquitin Binding Entity) agarose beads to detect Slp1Cdc20 ubiquitylation.

(F) TUBE pull-down assays in mts3-1 strains carrying sfGFP-tagged wild type or mutants of Slp1Cdc20 or pmk1Δ or pek1DD mutations. The TUBE beads-bound samples were blotted with anti-GFP antibodies. Asterisk indicates unspecific bands recognized by anti-GFP antibodies.

(G) Schematic summarizing the negative effect of the coupling of Pmk1 phosphorylation and K472/K479-mediated ubiquitylation on Slp1Cdc20 abundance and APC/C activation.

During the course of our mass spectrometry analysis on Slp1Cdc20 phosphorylation, we also detected Lys 479 (K479) as an ubiquitin-modifying site (Figure 4B, Figure 4-figure supplement 1 and Figure 4-figure supplement 2). We wondered whether K479 is involved in Slp1Cdc20 autoubiquitylation in fission yeast. In agreement with the mass spectrometry analysis, mutation of K479 to arginine indeed conferred stability of Slp1Cdc20 (Figure 5C), even in asynchronously growing cells (Figure 5-figure supplement 2), which is reminiscent of the stabilized Slp1Cdc20 in apc15Δ mutant cells at interphase (Sewart and Hauf, 2017). These results suggested that K479 is likely Slp1Cdc20 ubiquitylation-relevant. Furthermore, nda3-KM311 slp1K479R cells also demonstrated lowered metaphase arrest rate after cold treatment (Figure 5D and Figure 5-figure supplement 1), indicating defective SAC activation. Previous studies have identified two neighbouring lysine residues close to the carboxy terminus of human Cdc20, Lys485 and Lys490, as ubiquitylation sites in prometaphase cells, and mutating these two lysines to arginine completely prevented Cdc20 ubiquitylation (Danielsen et al., 2011; Mansfeld et al., 2011). Very interestingly, a second lysine residue, K472, is also present in fission yeast Slp1Cdc20 adjacent to K479 (Figure 4B and Figure 5-figure supplement 3). Although K472 was not identified as an ubiquitin-modifying residue in our mass spectrometry analysis, replacement of K472 with arginine also efficiently stabilized Slp1Cdc20, similar to K479R mutant (Figure 5C and Figure 5-figure supplement 2). More strikingly, simultaneously mutating both K472 and K479 caused additive effects in elevating Slp1Cdc20 levels and lowering SAC activation rate in wild-type or pek1DD strain background (Figure 5C, 5D, Figure 5-figure supplement 1 and Figure 5-figure supplement 2), suggesting that likely both these two lysine residues at the C-terminal tail of Slp1Cdc20 contribute to its turnover via ubiquitylation, which is very similar to the mechanism in human Cdc20. Consistently, using TUBEs (tandem ubiquitin-binding entities) (Hjerpe et al., 2009) to enrich for ubiquitinated proteins in metaphase-arrested mts3-1 cells expressing sfGFP-Slp1, we were able to demonstrate that mutating both K472 and K479 largely removed higher molecular weight bands detected by immunoblotting, which were corresponding to ubiquitylated species of Slp1Cdc20 (Figure 5E, F). We also noticed that Slp1K472R; K479R rendered more severely compromised viability than either single mutants (Figure 5-figure supplement 4), indicating blocking Slp1Cdc20 ubiquitylation could bring about detrimental consequence to the cells.

Our observation that Slp1Cdc20 levels were lowered in pek1DDmutant (Figure 2B) suggested that attenuation of Slp1Cdc20 levels upon constitutive activation of CIP signaling is very likely through facilitating its autoubiquitylation. To test this possibility, we examined and compared the ubiquitylation levels of sfGFP-Slp1 in wild type, pmk1Δ and pek1DD cells by affinity pull-down assays using TUBEs (Figure 5E). Indeed, we observed a much weaker or stronger discrete laddering pattern of sfGFP-Slp1 in pmk1Δ and pek1DD cells respectively than that of wild type cells, indicating abated or enhanced ubiquitylation (Figure 5F). Furthermore, mutations of T480A, S76A/T480A and Pmk1-docking site, but not S28A/T31A- or S28A/T31A/S59A-containing mutants, also reduced polyubiquitylation of Slp1Cdc20 in both pek1+ cells (Figure 5F) and pek1DD-overexpressing (i.e. Padh11-pek1DD) cells (Figure 5-figure supplement 5). All these results confirmed that Pmk1- but not Cdk1-mediated Slp1Cdc20 phosphorylation promotes its ubiquitylation.

Together, these findings lent further support to the idea that the CIP signaling pathway restrain APC/C activity through a cascade mechanism, in which Pmk1 directly binds and phosphorylates Slp1Cdc20 to facilitate its autoubiquitylation, and eventually lowers its levels (Figure 5G).

Osmotic stress and cell wall damage trigger rapid Slp1Cdc20 downregulation and mitotic exit delay

It has been well established that two MAPKs Sty1 and Pmk1 can be activated by environmental perturbations including osmotic stress, cell wall damage, thermal stress, oxidative stress, and glucose limitation, among others (Cansado et al., 2021; Madrid et al., 2006; Shiozaki and Russell, 1995b). We wondered whether the phosphorylation of Slp1Cdc20 by Pmk1 could tune the strength of SAC activation and APC/C activity in response to extracellular stress. To investigate this possibility, we performed arrest-and-release experiments with nda3-KM311 mutants and added 0.6 M KCl or 2 µg/mL of caspofungin 1 hr before release from metaphase arrest to elicit strong osmotic saline stress (KCl) or cell wall damage stress (caspofungin) response (Madrid et al., 2006; Madrid et al., 2016), respectively (Figure 6A). As previously reported, treatment with KCl or caspofungin resulted in full activation of both Pmk1 and Sty1 (for KCl treatment) or only Pmk1 (for caspofungin treatment) as determined using anti-phospho p42/44 or anti-phospho p38 antibodies, which serve as indicative of respective Pmk1 and Sty1 activation (Figure 6B). Intriguingly, treatment with KCl or caspofungin also correlated with a marked and rapid decrease in Slp1Cdc20 levels (Figure 6B). It is of note that SAC activation per se did not trigger activation of CIP pathway, as we did not observe any elevated Pmk1 phosphorylation in metaphase-arrested cells either by nda3-KM311 mutation (see “control” samples in Figure 6B) or by Mad2-overexpression (Figure 6-figure supplement 1).

Osmotic stress and cell wall damage trigger rapid Pmk1 phosphorylation-dependent Slp1Cdc20 downregulation and mitotic exit delay.

(A) Schematic depiction of the experimental design for treatment with KCl or caspofungin during nda3-mediated spindle checkpoint activation to activate MAPKs. Samples were collected at indicated time points for subsequent analyses including immunoblotting, co-IP and time-course analysis on SAC or APC/C activation.

(B) Immunoblot analysis of activation of MAPKs and Slp1Cdc20 protein levels. Samples with or without indicated treatments were blotted with anti-phospho p42/44 and anti-phospho p38 antibodies as indicative of phosphorylated Pmk1 (Pmk1-P) or phosphorylated Sty1 (Sty1-P), respectively. Slp1Cdc20 levels were detected with anti-Slp1 antibodies and anti-Cdc2 was used as loading control.

(C) Co-immunoprecipitation analysis of APC/C-MCC association upon environmental stress. Lid1-TAP was immunoprecipitated from nda3-KM311-arrested cells and associated Mad2-GFP, Mad3-GFP and Slp1Cdc20 were detected as in Figure 2A. Note that APC/C-MCC association was disrupted when 0.6 M KCl was present during cell culturing or during immunoprecipitation procedures.

(D) Time-course analyses of SAC activation and inactivation in nda3-KM311 cdc13-GFP strains with indicated genotypes after arrest at 18 °C and KCl or caspofungin treatments. The experiment was repeated 3 times and the mean value and p value for each sample were calculated as in Figure 1D.

(b) Time-course analyses of SAC activation and inactivation efficiency in Pmk1-docking- and phosphorylation-deficient slp1 mutants under environmental stresses elicited by 0.6 M KCl or 2 μg/mL caspofungin.

Next, we investigated whether environmental perturbations affect APC/C-MCC interaction. Surprisingly, we did not see any alteration of APC/C-MCC interaction accompanying Pmk1 activation when cells were treated by caspofungin (Figure 6C), which was different from what we observed in Padh11-pek1DD cells (Figure 2A). Also, we failed to detect enhanced APC/C-MCC association like we observed in Padh11-wis1DD cells when cells were treated by KCl, where Sty1 was activated (Figure 6C). The latter observation was most likely due to the disruptive effect of high concentration of KCl on protein-protein interactions, because we observed similar effect when we immunoprecipitated APC/C in the presence of 0.6 M KCl in the lysis buffer (Figure 6C, right panel). With this data, we could not exclude the possibility that activated Sty1 transiently enhanced APC/C-MCC association in yeast cells, which was soon counteracted by the disruptive effect of saline. Nevertheless, the high SAC activation efficiency in mitotis-arrested cells exposed to osmotic stress (KCl) or cell wall damage stress (caspofungin) was largely lowered by pmk1 deletion (Figure 6D and Figure 6-figure supplement 2), indicating that at least activated Pmk1 exerts regulation of the spindle checkpoint and APC/C activation under environmental stimuli. Further, Pmk1-docking site mutant slp1(DS-5A) and Pmk1 phosphorylation-deficient mutants slp1T480A and slp1S76A;T480A could partially suppress the sustained SAC activation responding to osmotic and cell wall damage stresses (Figure 6E and Figure 6-figure supplement 2). And accordingly, protein levels of Pmk1 phosphorylation-deficient mutants Slp1T480A and Slp1S76A;T480A under above stresses were restored almost to the wild-type Slp1 levels in untreated metaphase-arrested cells (Figure 6-figure supplement 3). We also noticed that protein levels of ubiquitylation-deficient mutants, including Slp1K472R, Slp1K479R and Slp1K472R;K479R, were more efficiently maintained than Slp1T480A and therefore more resistant to osmotic and cell wall damage stresses (Figure 6-figure supplement 3).

pmk1Δ cells are defective in faithful chromosome segregation upon environmental stress

Our above data demonstrated that, upon SAC activation, Pmk1 is involved in controlling the stability of Slp1Cdc20 and thus the activation of APC/C and anaphase entry under environmental stresses. We wondered whether the absence of pmk1+ would affect fidelity of chromosome segregation, particularly when cells are exposed to environmental stimuli. To answer this question, we treated G2-arrested and -released cdc25-22 cells with 0.6 M KCl or 2 μg/mL caspofungin to activate MAPKs (Figure 7A), and then examined possible chromosome segregation defects in pmk1Δ cells (Figure 7B). We found that stressed pmk1Δ cells displayed greatly increased frequency of lagging chromosomes and chromosome mis-segregation at mitotic anaphase compared to similarly treated wild-type cells and also untreated pmk1Δ cells (Figure 7C). This data further validated the importance of Pmk1 activation under adverse stress in delaying anaphase onset and maintaining genome integrity.

pmk1Δ cells are defective in faithful chromosome segregation upon environmental stress.

(A) Schematic depiction of the experiment design for arrest-and-release of cdc25-22 GFP-atb2+ strains. Environmental stresses were imposed by 0.6 M KCl or 2 μg/mL caspofungin. Samples were taken at 90 min after being shifted back from 36 °C to 25 °C to enrich anaphase cells

(B and C) Analyses of chromosome segregation in anaphase cells treated without or with KCl or caspofungin and after being fixed and stained with DAPI. Example pictures of anaphase cells with long spindles judged by GFP-Atb2 signals are shown (B). Anaphase cells with two categories of chromosome segregation defects (i.e. unequal chromosome segregation (mis-segregation) and lagging chromosomes) were quantified (C). >200 cells were counted for every sample. The mean values for each category were calculated, error bars indicate mean ± standard deviation of three independent experiments. p values were calculated with pooled data of two categories for each sample. n, numbers of anaphase cells analyzed. Scale bar, 5 μm.

Discussion

MAPKs are involved in regulation of mitotic anaphase onset in fission yeast

APC/C-mediated proteolysis of cyclin B and securin is key for anaphase entry by inactivating Cdk1 and permitting chromosome segregation, respectively. In response to unattached or tensionless kinetochores, the SAC induces MCC generation. As a critical APC/C inhibitor, MCC fulfills its function through the physical binding of BubR1/Mad3 and Mad2 to two molecules of Cdc20/Slp1Cdc20 (i.e. Cdc20MCC and Cdc20APC/C), and this mechanism has been confirmed at least in both humans and fission yeast (Alfieri et al., 2016; Chao et al., 2012; Izawa and Pines, 2015; Sewart and Hauf, 2017; Yamaguchi et al., 2016). Thus, Cdc20 is unique amongst the MCC proteins in that it is simultaneously a coactivator and inhibitor of the APC/C.

In the current study, we have uncovered a dual mechanism, in which two fission yeast MAPKs Pmk1 and Sty1 restrain APC/C activity through distinct manners, either by phosphorylating Slp1Cdc20 to lower its levels or by phosphorylating unknown substrate(s) to promote MCC-APC/C association and thus SAC signaling strength, respectively (Figure 8). Therefore, our study has established another critical APC/C-inhibitory mechanism in the spindle checkpoint, though it will be important for future studies to identify the phosphorylation target(s) posed by SAP signaling pathway. We should mention that most of our experimental setups in this study were based on cold-sensitive nda3-KM311 mutant, which causes loss of both kinetochore-microtubule attachments and intra-kinetochore tension when being grown at 18 °C. However, we were unable to distinguish whether MAPK involvement in SAC dynamics is relevant to one perturbation or another or both. Nevertheless, our current study together with earlier report (Edreira et al., 2020) have implicated fission yeast Pmk1 in both the mitotic and cytokinesis surveillance mechanisms in response to unfavorable environmental stimuli.

Summary of the mechanisms that how MAPKs negatively regulate APC/C activity in fission yeast.

(A) Schematic depiction of a possible dual mechanism that how activated MAPK signaling pathways are involved in delaying APC/C activation and SAC inactivation. Upon SAC and MAPK signaling activation, division of labor between the CIP and the SAP pathways enables the phosphorylation of Slp1Cdc20 and unidentified substrate(s) by Pmk1 and Sty1 respectively, which leads to lowered Slp1Cdc20 level and enhanced MCC affinity for APC/C.

(B) Summary of the alteration of Slp1Cdc20 protein levels and MCC-APC/C association strength under various conditions examined in this study. White wavy line indicates the absence of Apc15.

The regulation of mitotic entry by MAPK in fission yeast is triggered by many environmental insults, including high osmolarity, oxidative stress, heat shock, centrifugation, nutrient starvation and rapamycin treatment (Hartmuth and Petersen, 2009; Petersen and Hagan, 2005; Petersen and Nurse, 2007; Shiozaki and Russell, 1995a; Shiozaki et al., 1998). We found that treatment of fission yeast cells with KCl or caspofungin, which provokes a strong osmotic stress or cell wall damage response and quick Pmk1 activation, correlated with an attenuated Slp1Cdc20 protein levels and delayed APC/C activation. Thus, our finding implicates MAPK Pmk1 as an important regulator for anaphase entry in response to adverse extracellular conditions. Intriguingly, one previous study showed that treatment of human cells with the carcinogen cadmium activates MAPK p38 (fission yeast Sty1 homologue) and triggers accelerated ubiquitination and proteolysis of Cdc20, which is essential for prometaphase arrest under SAC (Yen and Yang, 2010). Therefore, although different MAPKs might be employed to fulfil the task in yeast and higher eukaryotes, it is still possible that phosphorylation and regulation of Cdc20 levels by MAPKs is an evolutionarily conserved mechanism to adjust mitotic exit to extracellular or environmental stimuli.

Negative phospho-regulation of Cdc20 by MAPK as a spindle assembly checkpoint-dependent APC/C-inhibitory mechanism

APC/C activity can be regulated at multiple levels, Cdk1-dependent APC/C phosphorylation is one of the major prerequisites for APC/C activation and anaphase onset, in which Apc3 and Apc1 are key subunits for positive phospho-regulation (Fujimitsu et al., 2016; Kraft et al., 2003; Qiao et al., 2016; Steen et al., 2008; Zhang et al., 2016). However, it has been shown that Cdk1-dependent phosphorylation of APC/C co-activator Cdc20 lessens its interaction with APC/C, thus serves as a negative phospho-regulation (Labit et al., 2012). In addition to its phosphorylation by Cdk1, phosphorylation of Cdc20 by multiple other kinases, such as Bub1, PKA, MAPK and Plk1, has also been demonstrated or suggested in Xenopus, humans or budding yeast to be inhibitory to APC/C activation and improper anaphase onset upon SAC activation ((Jia et al., 2016) and reviewed in (Yu, 2007)). Therefore, Cdc20 generally serves as an integrator of multiple intracellular signaling cascades that regulate progression through mitosis.

Although multiple kinases have been proposed to be directly involved in phosphorylation of Cdc20 proteins (Yu, 2007), precisely identifying the in vivo phosphorylation sites in Cdc20 has turned out to be challenging, largely due to its protein scarcity and instability. Based on known Cdk1 core consensus (SP/TP), 9 residues in human Cdc20 were originally suggested as putative Cdk1 sites, but among them only Ser41 has been proven to be phosphorylated in vivo by mass spectrometry analysis (Kramer et al., 2000; Yudkovsky et al., 2000). However, subsequent mass spectrometry analyses identified Ser41 as one of the Bub1 phosphorylation sites both in vivo and in vitro and another putative Cdk1 site Thr106 as one of the Plk1 sites (Jia et al., 2016; Tang et al., 2004). It is very likely that there are not as many genuine Cdk1 sites in Cdc20 as originally assumed. One supporting evidence is that mutation of only one (Thr32) of the three putative Cdk sites to alanine in N terminus of C. elegans CDC-20 is sufficient to phenocopy 3A mutant for accelerated APC/C activation (Kim et al., 2017). Interestingly, Thr32 of C. elegans CDC-20 is conserved as Thr 59 and Thr68 in human and Xenopus Cdc20, respectively (see Figure 4-figure supplement 4). Based on two-dimensional tryptic phosphopeptide analysis, Thr68 in Xenopus Cdc20 has also been regarded as a MAPK target site (Chung and Chen, 2003). These studies in diverse organisms raised an issue that whether multiple kinases share the same phosphorylation sites in Cdc20? It would not be so surprising if it turns out to be the case eventually, at least for Cdk1 and MAPKs, because CDKs and MAPKs recognize the same core motif (SP/TP) (Pinna and Ruzzene, 1996).

By purifying Apc15-aasociated fission yeast Slp1Cdc20 proteins from checkpoint-arrested pek1DD and pmk1Δ cells, we were able to identify Thr480 and Ser76 as in vivo Pmk1-relevant phosphorylation sites and three Pmk1-independent phosphorylation sites (S28, T31, S59) based on mass spectrometry analyses. All our identified sites except Thr480 are located in the N terminal disordered portion of Slp1Cdc20 sequence, this is similar to the phosphorylation residues identified by biochemical and functional analyses in other organisms, in which the major phosphorylation sites all fall within the N-terminus of Cdc20 homologues. We further provided evidence that S28/T31 and T480 are phosphorylated by Cdk1 and Pmk1, respectively, both in vitro and in vivo. We noticed that there is a tendency for reported phosphorylation residues to be concentrated in the N terminus of Cdc20 proteins (Jia et al., 2016; Tang et al., 2004; Yu, 2007). However, due to the lack of clear sequence homology in their N-terminal and C-terminal tails between Slp1Cdc20 and its Cdc20 homologues in higher eukaryotes (see Figure 4-figure supplement 4), it is impossible to certainly assign any identified sites from different organisms as homologous sites. But it is possible that multiple kinases modify shared sites in Cdc20, this collaborative effect may be especially helpful under adverse environmental stress. Nevertheless, all reported Cdc20 phosphorylation events constitute an APC/C-inhibitory mechanism, no matter what kinases are responsible for the modifications (Chung and Chen, 2003; D’Angiolella et al., 2003; Jia et al., 2016; Kim et al., 2017; Kramer et al., 2000; Labit et al., 2012; Tang et al., 2004; Yu, 2007; Yudkovsky et al., 2000).

Coupling of phosphorylation and autoubiquitylation of Slp1Cdc20

The current model of inhibitory signaling from Cdc20 posits that its phosphorylation either promotes the formation of MCC or inhibits APC/CCdc20 catalytically, both are required for proper spindle checkpoint signaling (Jia et al., 2016; Labit et al., 2012; Yamano, 2019). Interestingly, our current study suggests Slp1Cdc20 phosphorylation by fission yeast MAPK Pmk1 may represent another critical APC/C-inhibitory mechanism in the spindle checkpoint through reducing the protein levels of Slp1Cdc20, which is achieved via autoubiquitylation-mediated degradation (Figure 8). This mechanism is distinct from that reported in Xenopus egg extracts and tadpole cells, in which phosphorylation of Cdc20 by MAPK is instead responsible for facilitated Cdc20 binding by MCC to prevent it from activating the APC/C (Chung and Chen, 2003). Actually, previous studies in fission and budding yeast have underlined the importance of accurate relative abundance between checkpoint proteins and Cdc20/Slp1Cdc20, which sets an important determinant of checkpoint robustness (Heinrich et al., 2013; Pan and Chen, 2004). A benefit of allowing the association of MCC-bound Cdc20 with APC/C is to provide an opportunity for Cdc20 autoubiquitylation by APC/C, a process aided by APC/C subunit Apc15 (Mansfeld et al., 2011; May et al., 2017; Sewart and Hauf, 2017; Uzunova et al., 2012). Notably, the depletion of apc15+restores Slp1Cdc20 levels in pek1DD cells (Figure 5A), consistent with the idea that Pmk1-mediated Slp1Cdc20 phosphorylation very likely facilitates its autoubiquitylation. Indeed, our biochemical and functional analyses further establish K479 and K472 in Slp1Cdc20 as two key residues directly involved in its ubiquitylation, which likely follows its phosphorylation, particularly at adjacent Thr480. Despite K485 and K490 are similarly located at C terminus of human Cdc20 and can be ubiquitylated and responsible for Cdc20 degradation (Danielsen et al., 2011; Mansfeld et al., 2011), there has been no reports on phosphorylation-modified residues in this region. However, although the sequence homology within N- and C-terminal regions between Slp1Cdc20 and Cdc20 homologues in higher eukaryotes is extremely low, we do notice that there are several lysine-flanking serine or threonine residues within Cdc20 C-terminal tails of higher eukaryotes (see Figure 5-figure supplement 3), this raises a possibility that this region may be involved in phosphorylation-coupled ubiquitylation. Currently, we do not know how T480 phosphorylation accelerate ubiquitylation of K479 and K472 in Slp1Cdc20. It is possible that phosphorylation of Slp1Cdc20 might alter the mode of Cdc20 binding to APC/C which becomes conducive to catalyze Slp1Cdc20 ubiquitylation.

In summary, our studies revealed a distinct dual mechanism of MAPK-dependent anaphase onset delay imposed directly on APC/C co-activator Slp1Cdc20 and an unidentified substrate, which may emerge as an important regulatory layer to fine tune APC/C and MCC activity upon SAC activation. Importantly, although this mechanism of APC/C regulation by two MAPKs has not been examined in other organisms, it is very likely widespread across eukaryotes including humans, given the involvement of MAPKs in general response to wide range of adverse environmental situations and similar regulations of APC/C and SAC in eukaryotes characterized so far. It is also plausible to assume that the MAPK-mediated APC/C inhibition pathway may also contribute to the pathogenesis of multiple cancers. Indeed, Cdc20 upregulation has been demonstrated in several types of cancers, including breast and colon cancer, and it contributes to aggressive tumor progression and poor prognosis in gastric cancer and primary non-small cell lung cancer (Chi et al., 2019). One very recent study conducted in human pancreatic ductal adenocarcinoma (PDAC) cell lines revealed that inhibition of extracellular signal-regulated kinase (ERK) led to dephosphorylation of Cdc20 and concomitant loss of APC/C target proteins securin and cyclin B (CCNB1 and CCNB2) (Klomp et al., 2024). Because human ERKs are homologs of fission yeast Pmk1, the finding in this study supports a notion that very likely human MAPKs also negatively regulates APC/C through Cdc20 phosphorylation and the altered levels of Cdc20 in cancers are MAPK relevant. Nevertheless, inhibition of APC/CCdc20 activity or induction of Cdc20 degradation have already been listed as alternative therapeutic strategies to control cancer (Greil et al., 2022; Jeong et al., 2022). Thus, although the detailed molecular mechanisms may vary, further studies on MAPK-mediated APC/CCdc20 inhibition mechanism in human would lead to better understanding of its oncogenic role in tumor progression.

Materials and Methods

Fission yeast strains, media and genetic methods

Fission yeast cells were grown in either YE (yeast extract) rich medium or EMM (Edinburgh minimal medium) containing the necessary supplements. G418 disulfate (Sigma-Aldrich; A1720), hygromycin B (Sangon Biotech; A600230) or nourseothiricin (clonNAT; Werner BioAgents; CAS#96736-11-7) was used at a final concentration of 100 μg/ml and thiabendazole (TBZ) (Sigma-Aldrich; T8904) at 5-15 μg/ml in YE media where appropriate. For serial dilution spot assays, 10-fold dilutions of a mid-log-phase culture were plated on the indicated media and grown for 3 to 5 days at indicated temperatures.

To create strain with an ectopic copy of slp1+ at lys1+ locus on chromosome 1, the open reading frame of slp1+(1467 bp) and its upstream 1504 bp sequence was first cloned into the vector pUC119-Padh21-MCS-hphMX6-lys1* with adh21 promoter (Padh21) removed using the ‘T-type’ enzyme-free cloning method (Chen et al., 2017). The sequence of hphMX6 in above plasmid was replaced by fragment corresponding to nourseothiricin resistance gene sequence by BglII-EcoRI cut and re-ligation, generating pUC119-Pslp1-slp1+-Tadh1::natMX6-lys1*. The resultant plasmids were linearized by ApaI and integrated into the lys1+locus, generating the strains lys1Δ::Pslp1-slp1+-Tadh1::hphMX6 and lys1Δ::Pslp1-slp1+-Tadh1::natR.

To generate the mutant strains harboring lysine/arginine (K/R) or serine/threonine to glutamic acid (E) or alanine (A) mutations, or lysine (K) to arginine (R) mutations within Slp1, the desired mutations were introduced into pUC119-Pslp1-slp1+-Tadh1::hphMX6 using standard methods of site-directed mutagenesis. To generate the strains carrying superfolder GFP (sfGFP)-tagged wild type or mutant Slp1, a fragment corresponding to the sequence of sfGFP was inserted in front of slp1 in pUC119-Pslp1-slp1-Tadh1::hphMX6-lys1* using the ‘T-type’ enzyme-free cloning method. The resulting plasmids were linearized and integrated into the lys1+ locus of chromosome 1 using the hygr marker. Then endogenous slp1+ was deleted and replaced with ura4+.

To generate the strains expressing ade6::Padh1-GST-slp1(456-488aa)::natR, C-terminal sequence of slp1+ corresponding to 456-488aa was first cloned into the vector pUC119-Padh1-MCS-natMX6-lys1* and the sequence of lys1* was replaced with ade6+ from pAde6-NotI-hphMX (a kind gift of Li-lin Du), and then the sequence of GST from pGEX-4T-1 was inserted in front of slp1. The resulting plasmid pUC119-Padh1-GST-slp1(456-488aa)-natMX6-ade6 was linearized and integrated into the ade6+locus of chromosome III using the natr marker.

To generate the strains expressing ade6::Padh1-slp1(1-60aa)-mEGFP-2xNLS-GST-slp1(456-488aa)::natR, sequences corresponding to slp1(1-60aa), mEGFP and two tandem SV40 NLS (CCT AAG AAA AAA CGA AAA GTT GAG GAT CCT AAA AAG AAA CGA AAA GTT GAT) were cloned into the plasmid pUC119-Padh1-GST-slp1(456-488aa)-natMX6-ade6 using the ‘T-type’ enzyme-free cloning method. The resultant plasmid was linearized and integrated into the ade6+locus as describes above.

To generate the strains expressing lys1Δ::Padh11-pmk1-GBP-mCherry::hygR, the coding sequence of pmk1+ was first cloned into the vector pUC119-Padh11-GBP-mCherry-hphMX6-lys1* (Chen et al., 2017), then the plasmid was linearized and integrated into the lys1+ locus in the genome.

To generate the constitutively active MAPKK strains pek1DDor wis1DD, the Ser234 and Thr238 of pek1+ in pUC119-Padh21-6HA-pek1+-Tadh1-hphMX6-lys1*, pUC119-Padh11-6HA-pek1+-Tadh1-hphMX6-lys1* and pUC119-Padh11-6HA-pek1+-Tadh1-kanMX6-ura4+ and the Ser469 and Thr473 of wis1+in pUC119-Padh21-6HA-wis1+-Tadh1-hphMX6-lys1*, pUC119-Padh11-6HA-wis1+-Tadh1-hphMX6-lys1* and pUC119-Padh11-6HA-wis1+-Tadh1-kanMX6-ura4+were changed to aspartic acids (D) by standard methods of site-directed mutagenesis. The resulting plasmids were linearized and integrated into the lys1+ locus of chromosome I or the ura4+locus of chromosome III using the kanr marker. A list of the yeast strains used is in Supplementary file 1.

Fission yeast cell synchronization methods

For nda3-KM311 strains, cells were grown at the permissive temperature for nda3-KM311 (30 °C) to mid-log phase, synchronized at S phase by adding HU (Sangon Biotech; A600528) to a final concentration of 12 mM for 2 hours followed by a second dose of HU (6 mM final concentration) for 3.5 hours. HU was then washed out and cells were released at specific temperatures as required by subsequent experiments.

For cdc25-22 background strains, cells were first grown at 25 °C and then arrested at the G2/M transition by shifting to 36 °C for 3.5 hours. To allow cells to progress into mitosis, the cultures were released by shifting back to 25 °C. Progression through cell cycle was monitored by DAPI staining after being fixed and counting binucleate cells.

Spindle checkpoint activation and silencing assays in fission yeast

For checkpoint silencing assay in the absence of microtubules, mid-log cdc13-GFP nda3-KM311 cells were first synchronized with HU at 30 °C and then arrested in early mitosis by shifting to 18 °C for 6 hours, followed by incubation at 30 °C to allow spindle reformation and therefore spindle checkpoint inactivation. For osmotic stress or cell wall damage stress experiments, 0.6 M KCl or 2 μg/mL caspofungin was added into cultures during 18 °C incubation 1 hr before being shifted to 30 °C. Cells were withdrawn at certain time intervals and fixed with cold methanol and stained with DAPI. 200-300 cells were analyzed for each time point. Each experiment was repeated at least three times.

To achieve spindle checkpoint activation and metaphase arrest through a different mechanism, Mad2 was overexpressed from the nmt1 promotor (i.e. Pnmt1). To induce Mad2 overexpression, strains carrying Pnmt1-mad2::leu1+were pre-cultured in minimal medium with supplements (EMM5S) and 15 µM thiamine, cells were washed 5 times in EMM5S to remove thiamine before being grown in EMM5S at 30 °C for 18 hours. Mitotic arrest was determined by DAPI staining and GFP-Atb2-labeled short spindles.

Immunoblotting and immunoprecipitation

For routine Western blot and immunoprecipitation experiments, yeast cells were collected from nda3-KM311 arrested and released cultures, followed by lysing with glass bead disruption using Bioprep-24 homogenizer (ALLSHENG Instruments, Hangzhou, China) in NP40 lysis buffer (6 mM Na2HPO4, 4 mM NaH2PO4, 1% NP-40, 150 mM NaCl, 2 mM EDTA, 50 mM NaF, 0.1 mM Na3VO4) plus protease inhibitors as previously described (Wang et al., 2012). Proteins were immunoprecipitated by IgG Sepharose beads (GE Healthcare; 17-0969-01) (for Lid1-TAP) or anti-GFP nanobody agarose beads (AlpalifeBio, Shenzhen, China; KTSM1301) (for Mad3-GFP). Immunoblot analysis of cell lysates and immunoprecipitates was performed using appropriate antibodies at 1:500 - 1:5000 dilutions and was read out using chemiluminescence.

Detection of in vivo phosphorylation of Slp1Cdc20 at Thr480 and Ser28/Thr31

For detection of phosphorylation of Slp1Cdc20 at Thr480 in vivo, mts3-1 mutant strains expressing ade6::Padh1-GST-slp1(456-488aa)::natR in wild type, pmk1Δ and lys1Δ::Padh11-6xHA-pek1DDbackground were arrested at 36 [for 3.5 hours, followed by pull-down of GST-Slp1(456-488aa) by glutathione Sepharose 4B (GE Healthcare). Purified GST-Slp1(456-488aa) was detected with either goat anti-GST HRP-conjugated antibodies (GE Healthcare) or rabbit polyclonal anti-pThr480 antibodies.

For detection of phosphorylation of Slp1Cdc20 at Ser28/Thr31 in vivo, nda3-KM311 apc15-13myc::kanR strains expressing slp1+ or slp1(S28A;T31A) in wild type or cdc2-asM17 background were first synchronized with HU at 30 °C and then arrested in early mitosis by shifting to 18 °C for 6 hours. For cdc2-asM17 strains, 1 μM 1-NM-PP1 (Toronto Research Chemicals; A603003) was added into one of the cultures to inactivate Cdc2 (Cdk1) 20 min before being collected. Apc15-13myc was immunoprecipitated by anti-c-Myc magnetic beads (MedChemExpress (MCE); HY-K0206). To remove Slp1Cdc20 phosphorylation, one of the immunoprecipitated samples was treated with 400 units of Lambda protein phosphatase (New England Biolabs, P0753S) at 30 °C for 30 min. Co-immunoprecipitated Slp1Cdc20 was detected with rabbit polyclonal anti-Slp1 antibodies, and the phosphorylated form of Slp1Cdc20 was detected with rabbit polyclonal anti-pSer28/pThr31 antibodies respectively.

Purification of APC/C and analyses by mass spectrometry

To prepare APC/C for mass spectrometry analyses, Apc15-GFP was purified from 6-liter cultures of nda3-KM311 arrested cells with desired genetic background. Cells were disrupted and cell lysates were prepared as described above for routine immunoprecipitation experiments.

For mass spectrometry analyses, purified samples were first run on PAGE gels, after staining of gels with Coomassie blue, excised gel segments were subjected to in-gel trypsin (Promega, V5111) digestion and dried. Samples were then analyzed on a nanoElute (plug-in V1.1.0.27; Bruker, Bremen, Germany) coupled to a timsTOF Pro (Bruker, Bremen, Germany) equipped with a CaptiveSpray source. Peptides were separated on a 15cm X 75μm analytical column, 1.6 μm C18 beads with a packed emitter tip (IonOpticks, Australia). The column temperature was maintained at 55 °C using an integrated column oven (Sonation GmbH, Germany). The column was equilibrated using 4 column volumes before loading sample in 100% buffer A (99.9% MilliQ water, 0.1% FA) (Both steps performed at 980 bar). Samples were separated at 400 nL/min using a linear gradient from 2% to 25% buffer B (99.9% ACN, 0.1% FA) over 90 min before ramping to 37% buffer B (10min), ramp to 80% buffer B (10min) and sustained for 10 min (total separation method time 120 min). The timsTOF Pro (Bruker, Bremen, Germany) was operated in PASEF mode using Compass Hystar 6.0. Mass Range 100 to 1700m/z, 1/K0 Start 0.6 V⋅s/cm2 End 1.6 V⋅s/cm2, Ramp time 110.1ms, Lock Duty Cycle to 100%, Capillary Voltage 1600V, Dry Gas 3 L/min, Dry Temp 180°C, PASEF settings: 10 MS/MS scans (total cycle time 1.27sec), charge range 0-5, active exclusion for 0.4 min, Scheduling Target intensity 10000, Intensity threshold 2500, CID collision energy 42eV. All raw files were analyzed by PEAKS Studio Xpro software (Bioinformatics Solutions Inc., Waterloo, ON, Canada). Data was searched against the S. pombe proteome sequence database (Uniprot database with 5117 entries of protein sequences at https://www.uniprot.org/proteomes/UP000002485). De novo sequencing of peptides, database search and characterizing specific PTMs were used to analyze the raw data; false discovery rate (FDR) was set to ≤1%, and [-10*log(p)] was calculated accordingly where p is the probability that an observed match is a random event. The PEAKS used the following parameters: (i) precursor ion mass tolerance, 20 ppm; (ii) fragment ion mass tolerance, 0.05 Da (the error tolerance); (iii) tryptic enzyme specificity with two missed cleavages allowed; (iv) monoisotopic precursor mass and fragment ion mass; (v) a fixed modification of cysteine carbamidomethylation; and (vi) variable modifications including N-acetylation of proteins and oxidation of Met.

Ubiquitin affinity pull-down assays

For detection of in vivo ubiquitylation of Slp1Cdc20, strains all carried mts3-1 mutation were used. sfGFP-tagged wild type Slp1 or Slp1 mutants were expressed in the genomically integrated form of Pslp1-sfGFP-slp1(WT or mutants)::hygR in wild type, pmk1Δ or Padh11-6xHA-pek1DDbackground. Cultures were first grown at 25 °C to mid-log phase and then shifted to 36 [for 3.5 hours to block cells in mitosis prior to harvesting and cell lysis. Ubiquitinated proteins were pulled down from yeast lysates using Tandem Ubiquitin Binding Entities (TUBEs) (LifeSensors, UM402) according to the manufacturer’s instructions. The bound proteins were eluted from washed Agarose-TUBE2 beads and analyzed by western blotting using a mouse monoclonal anti-GFP antibody (Beijing Ray Antibody Biotech; RM1008). Cell extracts were also immunoblotted for GFP and Cdc2 as loading controls (inputs).

Expression and purification of recombinant proteins

GST or MBP fusion constructs of fission yeast full-length or truncated Slp1Cdc20, Pmk1, Pek1 and Atf1 were generated by PCR of the corresponding gene fragments from yeast genomic DNA or cDNA and cloned in frame into expression vectors pGEX-4T-1 (GE Healthcare) or pMAL-c2x (New England Biolabs), respectively. Site-directed mutagenesis was done by standard methods to generate vectors carrying Pek1-DD(S234D,[T238D) or Slp1Cdc20 harboring K(Lys)/R(Arg) to E(Glu) or A(Ala) mutations at individual or combined sites including K19, K20, R21, K47, R48, K140, K141, K161, R162, R163, K472, R473. Integrity of cloned DNA was verified by sequencing analysis. All recombinant GST- or MBP-fusion proteins were expressed in Escherichia coli BL21 (DE3) cells. The recombinant proteins were purified by glutathione Sepharose 4B (GE Healthcare) or Amylose Resin High Flow (New England Biolabs) respectively according to the manufacturers’ instructions.

In vitro pulldown assay

The recombinant MBP-Slp1Cdc20 and MBP, or GST-Pmk1 and GST proteins were produced in bacteria and purified as described above. To detect the interaction between bacterially expressed Slp1Cdc20 and Pmk1-HA expressed in yeast, about 1 μg of MBP-Slp1Cdc20 or MBP immobilized on amylose resin was incubated with cleared cell lysates prepared from yeast strains nda3-KM311 pmk1-6His-HA, which were HU synchronized and arrested at prometaphase by being grown at 18 °C for 6 hours. To detect the direct interaction between bacterially expressed Slp1 and Pmk1, about 1 μg of GST-Pmk1 or GST immobilized on glutathione Sepharose 4B resin was incubated with cleared bacterial lysates expressing MBP-Slp1Cdc20. Samples were incubated for 1-3 hours at 4 °C. Resins were washed 3 times with lysis buffer (6 mM Na2HPO4, 4 mM NaH2PO4, 1% NP-40, 150 mM NaCl, 2 mM EDTA, 50 mM NaF, 0.1 mM Na3VO4, plus protease inhibitors), suspended in SDS sample buffer, and then subject to SDS-PAGE electrophoresis and Coomassie brilliant blue (CBB) staining.

Non-radioactive in vitro kinase assay

GST-Slp1Cdc20, GST-Atf1, GST-Pmk1, and GST-Pek1DD fusions were expressed and purified from Escherichia coli with glutathione Sepharose 4B as described above. Pmk1-HA-His and Cdc13-GFP were purified from yeast lysates by immuprecipitation with anti-HA or anti-GFP antibodies respectively.

For detection of phosphorylation of Slp1Cdc20 and Atf1 by yeast derived Pmk1, glutathione Sepharose 4B beads bound with GST-Slp1Cdc20 or GST-Atf1 and agarose beads bound with Pmk1-HA-His were mixed and washed 3 times with kinase buffer (50 mM Tris-HCl, pH 7.5, 10 mM MgCl2, 1 mM EGTA), then the reactions were incubated in kinase buffer with 20[μM ATPγS (Sigma-Aldrich; A1388) at 30°C for 45[min. The kinase reactions were stopped by adding 20[mM EDTA, and the reaction mixture was alkylated after incubation at room temperature with 2.5[mM p-nitrobenzyl mesylate (PNBM) (Abcam; ab138910) for 1 hr. The alkylation reaction was stopped by boiling in SDS-PAGE loading buffer. Phosphorylated proteins were detected with rabbit monoclonal anti-Thiophosphate ester antibody (Abcam, ab239919).

For detection of phosphorylation of Slp1Cdc20 at Thr480 by bacterially expressed Pmk1, beads bound with GST-Pmk1, GST-Pek1(DD) and fused substrates GST-Slp1(456-488aa) or GST-Slp1(456-488aa, T480A) were washed and incubated in kinase buffer (50 mM Tris-HCl, pH 7.5, 10 mM MgCl2, 1 mM EGTA) with 10 mM ATP (Sangon Biotech; A600020-0005) at 30°C for 45[min. The reaction was stopped by boiling in SDS-PAGE loading buffer. Phosphorylated proteins were detected with rabbit polyclonal anti-pThr480 antibodies.

For detection of phosphorylation of Slp1Cdc20 at Ser28/Thr31 by Cdk1, beads bound with bacterially expressed MBP-Slp1(1-190aa) or MBP-Slp1(1-190aa,S28A;T31A) and agarose beads bound with Cdc13-GFP/Cdc2 from yeast lysate were mixed and washed 3 times with kinase buffer (10 mM Tris-HCl, pH 7.4, 10 mM MgCl2, 1 mM DTT), then the reactions were incubated in kinase buffer with 10 mM ATP (Sangon Biotech; A600020-0005) at 30°C for 45[min. To inactivate Cdc2-as1, 5 μM 1-NM-PP1 (Toronto Research Chemicals; A603003) was added into the reaction. The reaction was stopped by boiling in SDS-PAGE loading buffer. Phosphorylated proteins were detected with rabbit polyclonal anti-pSer28/pThr31 antibodies.

Antibodies for immunoblotting

The phospho-site specific antibodies were made in an in-house facility (for anti-Slp1-pS28/pT31) or by ABclonal Technology Co., Ltd. (Wuhan, China) (for anti-Slp1-pT480) by immunizing rabbits with pS28/pT31- or pT480-containing peptides (LVFPN(S-p)PI(T-p)PLHQQ or PITK(T-p)PSSS, respectively) coupled to haemocyanin (Sigma). The purified antibodies were concentrated to 0.2 mg/mL and used at 1:100 dilution. The following antibodies used for immunoblot analyses were purchased from the indicated commercial sources and were used at the indicated dilution: peroxidase-anti-peroxidase (PAP) soluble complex (Sigma-Aldrich; P1291; 1:10,000); rabbit polyclonal anti-Myc (GeneScript; A00172-40; 1:2,000); mouse monoclonal anti-GFP (Beijing Ray Antibody Biotech; RM1008; 1:2,000); rat monoclonal anti-HA (Roche, Cat. No. 11 867 423 001; 1:2,000); rabbit polyclonal anti-Slp1 (Kim et al., 1998) (1:500); rabbit polyclonal anti-phospho-p44/42 (detecting activated Pmk1) (Cell Signaling Technology; #9101; 1:1,000); rabbit monoclonal anti-phospho-p38 (Cell Signaling Technology; #4511; 1:1,000); rabbit polyclonal anti-PSTAIRE (detecting Cdc2) (Santa Cruz Biotechnology; sc-53; 1:1,000); rabbit monoclonal anti-Thiophosphate ester antibody (Abcam; ab239919; 1:5,000); goat anti-GST HRP-conjugated antibody (RRID:AB_771429; GE Healthcare; 1:10,000). Secondary antibodies used were goat anti-mouse or goat anti-rabbit polyclonal IgG (H+L) HRP conjugates (Thermo Fisher Scientific; #31430 or #32460; 1:10,000).

RT-qPCR

For RT-qPCR of slp1+ mRNA, nda3-KM311 strains were grown, synchronized by HU and arrested at metaphase with SAC being activated as described above. Cells were collected and total cellular RNA was isolated by using TRIZOL method. cDNA was prepared using oligo(dT) and PrimeScript reverse transcriptase (Takara, D2680A). Real-time PCR was performed in the presence of SYBR Green using an Applied Biosystems 7900HT light cycler. Relative RNA levels were calculated using the ΔCT method and normalized to act1+ levels.

Fluorescence microscopy

Cdc13-GFP proteins were observed in cells after fixation with cold methanol. For DAPI staining of nuclei, cells were fixed with cold methanol, washed in PBS and resuspended in PBS plus 1 μg/ml DAPI. Photomicrographs of cells were obtained using a Nikon 80i fluorescence microscope coupled to a cooled CCD camera (Hamamatsu, ORCA-ER) or a Perkin Elmer spinning-disk confocal microscope (UltraVIEW® VoX) with a 100x NA 1.49 TIRF oil immersion objective (Nikon) coupled to a cooled CCD camera (9100-50 EMCCD; Hamamatsu Photonics) and spinning disk head (CSU-X1, Yokogawa). Image processing and analysis were carried out using Element software (Nikon), ImageJ software (National Institutes of Health) and Adobe Photoshop.

Statistical analyses and reproducibility

No statistical methods were used to predetermine sample size. Sample size was as large as practicable for all experiments. The experiments were randomized and the investigators were blinded to group allocation during experiments but not blinded to outcome assessment and data analysis.

All experiments were independently repeated two to more than three times with similar results obtained. For quantitative analyses of each time course experiment, ≥300 cells were counted for each time point or sample. The same sample was not measured or counted repeatedly. No data were excluded from our studies. Data collection and statistical analyses were performed using Microsoft Office Excel or GraphPad Prism 6 softwares. Data are expressed as mean values with error bars (±[standard deviation/s.d.) and were compared using two-tailed Student’s t-tests unless indicated otherwise.

Supporting information